---
title: "benchmarking"
output: html_document
date: "2026-03-29"
---

```{r}
library(tidyverse)
library(SpatialExperiment)
library(spicyR)
library(TreeBH)
library(DART)
library(purrr)
library(dplyr)

source("R/globals.R")
source("R/simulation_helpers.R")
source("R/benchmarking_helpers.R")
source("R/pfilter.R")
```


```{r}
# Core simulation outputs
single_df    <- readRDS("results/simulation/single_df.rds")
repeated_df  <- readRDS("results/simulation/repeated_df.rds")

# Sensitivity analysis simulation outputs
single_sensitivity_dfs   <- readRDS("results/simulation/single_sensitivity_dfs.rds")
repeated_sensitivity_dfs <- readRDS("results/simulation/repeated_sensitivity_dfs.rds")

# spicy outputs
spicy_single_sub     <- readRDS("results/spicy/spicy_single_sub.rds")
spicy_single_major   <- readRDS("results/spicy/spicy_single_major.rds")
spicy_repeated_sub   <- readRDS("results/spicy/spicy_repeated_sub.rds")
spicy_repeated_major <- readRDS("results/spicy/spicy_repeated_major.rds")

# Sensitivity analysis spicy outputs
single_sensitivity_spicy   <- readRDS("results/spicy/single_sensitivity_spicy.rds")
repeated_sensitivity_spicy <- readRDS("results/spicy/repeated_sensitivity_spicy.rds")
```

```{r}
topPairs(spicy_single_sub)
topPairs(spicy_single_major)

topPairs(spicy_repeated_sub)
topPairs(spicy_repeated_major)

spicy_single_sub$p.value["CD4_T__NK",]
spicy_single_sub$p.value["CD8_T__NK",]
spicy_single_sub$p.value["Memory_T__NK",]

summary(spicy_single_sub)
head(rownames(spicy_single_sub$p.value), 20)
spicy_single_sub$p.value["CD4_T__NK",]

"CD4_T__NK" %in% rownames(spicy_single_sub$p.value)
"CD8_T__NK" %in% rownames(spicy_single_sub$p.value)
"Memory_T__NK" %in% rownames(spicy_single_sub$p.value)

extract_leaf_stats(spicy_single_sub) %>% 
  filter(leaf %in% c("CD4_T__NK", "CD8_T__NK", "Memory_T__NK"))

# Total pairs
nrow(spicy_single_sub$p.value)

# Self pairs (A__A)
sum(sapply(rownames(spicy_single_sub$p.value), function(x) {
  ab <- strsplit(x, "__", fixed = TRUE)[[1]]
  ab[1] == ab[2]
}))
extract_leaf_stats(spicy_single_sub) %>% nrow()
```


# Ground Truth

The simulated signal is co-localisation between T cell subtypes and NK cells.
The three true positive pairs (in canonical `A__B` form) are defined here and
used throughout all benchmarking computations.

```{r ground_truth}
TRUE_PAIRS <- c("CD4_T__NK", "CD8_T__NK", "Memory_T__NK")
```

# Benchmarking Helpers

```{r helpers}
#' Compute TPR, FDR, FPR, Precision, and F1 from a results tibble.
#'
#' @param result_df  A tibble with at least columns `leaf`, `reject`, `is_true`.
#' @param true_pairs Character vector of true positive canonical pair names.
#'
#' @return A one-row tibble of performance metrics.
compute_metrics <- function(result_df, true_pairs = TRUE_PAIRS) {
  result_df <- result_df %>%
    mutate(
      is_true  = leaf %in% true_pairs,
      rejected = as.logical(reject)
    )

  TP <- sum( result_df$is_true  &  result_df$rejected)
  FP <- sum(!result_df$is_true  &  result_df$rejected)
  FN <- sum( result_df$is_true  & !result_df$rejected)
  TN <- sum(!result_df$is_true  & !result_df$rejected)

  TPR       <- if ((TP + FN) > 0) TP / (TP + FN) else NA_real_
  FDR       <- if ((TP + FP) > 0) FP / (TP + FP) else 0
  FPR       <- if ((FP + TN) > 0) FP / (FP + TN) else NA_real_
  Precision <- if ((TP + FP) > 0) TP / (TP + FP) else NA_real_
  F1        <- if (!is.na(Precision) && !is.na(TPR) && (Precision + TPR) > 0)
                 2 * Precision * TPR / (Precision + TPR) else NA_real_

  tibble(TPR, FDR, FPR, Precision, F1, TP, FP, FN, TN)
}

#' Run all methods on a pair of spicy sub/major results.
#'
#' @param spicy_sub   spicyR result at the sub-type level.
#' @param spicy_major spicyR result at the major-type level.
#' @param spe         SpatialExperiment object (needed for DART).
#' @param true_pairs  Character vector of ground-truth canonical pair names.
#'
#' @return A list with two elements:
#'   \item{raw}{Tidy tibble, one row per method x leaf, with columns
#'     method, leaf, pval, coef (where available), reject, is_true.}
#'   \item{metrics}{Tidy tibble, one row per method, with columns
#'     method, TPR, FDR, FPR, Precision, F1, TP, FP, FN, TN.}
run_all_methods <- function(spicy_sub, spicy_major, spe,
                            true_pairs = TRUE_PAIRS) {

  # ── Flat BH (baseline) ───────────────────────────────────────────────
  flat_bh <- extract_leaf_stats(spicy_sub) %>%
    mutate(method = "Flat BH", reject = bh_reject(pval, q=0.05))

  # ── Flat BY (baseline under arbitrary dependence) ────────────────────
  flat_by <- extract_leaf_stats(spicy_sub) %>%
    mutate(
      method = "Flat BY",
      reject = p.adjust(pval, method = "BY") <= 0.05
    )

  # ── TreeBH ───────────────────────────────────────────────────────────
  treebh <- tryCatch(
    run_treebh(spicy_sub),
    error = function(e) {
      message("TreeBH failed: ", conditionMessage(e))
      extract_leaf_stats(spicy_sub) %>% mutate(method = "TreeBH", reject = FALSE)
    }
  )

  # ── DART ─────────────────────────────────────────────────────────────
  dart <- tryCatch(
    run_dart_one(spicy_sub, spe),
    error = function(e) {
      message("DART failed: ", conditionMessage(e))
      extract_leaf_stats(spicy_sub) %>% mutate(method = "DART", reject = FALSE)
    }
  )

  # ── p-filter ─────────────────────────────────────────────────────────
  pfilter_res <- tryCatch(
    run_pfilter_one(spicy_sub, spicy_major),
    error = function(e) {
      message("pfilter failed: ", conditionMessage(e))
      extract_leaf_stats(spicy_sub) %>% mutate(method = "pfilter", reject = FALSE)
    }
  )

  # ── Ours (level-wise BH) ─────────────────────────────────────────────
  ours_res <- tryCatch({
    out <- run_ours_one(spicy_major, spicy_sub)
    out$res_ours %>%
      filter(level == 2) %>%
      transmute(leaf = node, pval = pvalue, method = "Ours", reject)
  }, error = function(e) {
    message("Ours failed: ", conditionMessage(e))
    extract_leaf_stats(spicy_sub) %>% mutate(method = "Ours", reject = FALSE)
  })

  # ── ClusterTrim (BY) ─────────────────────────────────────────────────
  clustertrim <- tryCatch(
    run_cluster_trim_one(spicy_sub, spicy_major),
    error = function(e) {
      message("ClusterTrim_BY failed: ", conditionMessage(e))
      extract_leaf_stats(spicy_sub) %>%
        mutate(method = "ClusterTrim_BY", reject = FALSE)
    }
  )

  # ── Combine into raw tibble ──────────────────────────────────────────
  all_raw <- list(flat_bh, flat_by, treebh, dart, pfilter_res, ours_res, clustertrim) %>%
    purrr::map_dfr(function(df) {
      df %>%
        select(any_of(c("method", "leaf", "pval", "coef", "reject"))) %>%
        mutate(
          reject  = as.logical(reject),
          is_true = leaf %in% true_pairs
        )
    })

  # ── Compute per-method metrics from raw ──────────────────────────────
  metrics <- all_raw %>%
    group_by(method) %>%
    group_modify(~ compute_metrics(.x, true_pairs)) %>%
    ungroup()

  list(raw = all_raw, metrics = metrics)
}

#' Display the leaf-level raw results as a formatted kable table.
#'
#' @param raw_df  The `$raw` tibble from `run_all_methods()`, optionally with
#'                extra columns (signal_weight, scenario) already attached.
#' @param caption Table caption string.
display_raw_results <- function(raw_df, caption = "Leaf-level rejection results") {
  raw_df %>%
    mutate(
      pval    = signif(pval, 3),
      coef    = if ("coef" %in% names(.)) signif(coef, 3) else NA_real_,
      reject  = ifelse(reject, "Yes", "No"),
      is_true = ifelse(is_true, "\u2713", "")
    ) %>%
    select(any_of(c("method", "leaf", "pval", "coef",
                    "reject", "is_true", "signal_weight", "scenario"))) %>%
    arrange(method, leaf) %>%
    filter(reject == "Yes")
}
```

# Core Benchmarking

## Single Measure (20 Subjects)

```{r benchmark_single}
spe_single     <- make_spe_from_df(single_df)
results_single <- run_all_methods(spicy_single_sub, spicy_single_major, spe_single)

raw_single     <- results_single$raw     %>% mutate(scenario = "Single measure")
metrics_single <- results_single$metrics %>% mutate(scenario = "Single measure")

saveRDS(raw_single,     "results/simulation/raw_single.rds")
saveRDS(metrics_single, "results/simulation/metrics_single.rds")
```

### Leaf-level Results — Single Measure

```{r raw_single_table}
display_raw_results(
  raw_single,
  caption = "Single Measure — per-leaf rejection decisions (\u2713 = true signal pair)"
)
```

### Performance Metrics — Single Measure

```{r metrics_single_table}
metrics_single %>%
  select(method, TPR, FDR, FPR, Precision, F1) %>%
  arrange(desc(TPR))
```

## Repeated Measures (5 Subjects × 5 Images)

```{r benchmark_repeated}
spe_repeated     <- make_spe_from_df(repeated_df)
results_repeated <- run_all_methods(spicy_repeated_sub, spicy_repeated_major, spe_repeated)

raw_repeated     <- results_repeated$raw     %>% mutate(scenario = "Repeated measures")
metrics_repeated <- results_repeated$metrics %>% mutate(scenario = "Repeated measures")

saveRDS(raw_repeated,     "results/simulation/raw_repeated.rds")
saveRDS(metrics_repeated, "results/simulation/metrics_repeated.rds")
```

### Leaf-level Results — Repeated Measures

```{r raw_repeated_table}
display_raw_results(
  raw_repeated,
  caption = "Repeated Measures — per-leaf rejection decisions (\u2713 = true signal pair)"
)
```

### Performance Metrics — Repeated Measures

```{r metrics_repeated_table}
metrics_repeated %>%
  select(method, TPR, FDR, FPR, Precision, F1) %>%
  arrange(desc(TPR)) %>%
  knitr::kable(digits = 3, caption = "Repeated Measures — Method Performance")
```

## Combined Summary Table

```{r combined_table}
bind_rows(metrics_single, metrics_repeated) %>%
  select(scenario, method, TPR, FDR, FPR, Precision, F1) %>%
  arrange(scenario, desc(TPR)) %>%
  knitr::kable(digits = 3, caption = "All Scenarios — Method Performance Summary")
```

# Sensitivity Analysis

## Single Measure — Run All Methods Across Signal Weights

```{r sensitivity_single_benchmark}
single_sensitivity_results <- purrr::imap(
  single_sensitivity_dfs,
  function(df, sw_name) {
    sw <- as.numeric(sub("sw_", "", sw_name))
    message(sprintf("Benchmarking single measure: %s", sw_name))

    spe         <- make_spe_from_df(df)
    spicy_sub   <- single_sensitivity_spicy[[sw_name]]$sub
    spicy_major <- single_sensitivity_spicy[[sw_name]]$major

    out <- run_all_methods(spicy_sub, spicy_major, spe)
    list(
      raw     = out$raw     %>% mutate(signal_weight = sw, scenario = "Single measure"),
      metrics = out$metrics %>% mutate(signal_weight = sw, scenario = "Single measure")
    )
  }
)

single_sensitivity_raw     <- purrr::map_dfr(single_sensitivity_results, "raw")
single_sensitivity_metrics <- purrr::map_dfr(single_sensitivity_results, "metrics")

saveRDS(single_sensitivity_raw,     "results/spicy/single_sensitivity_raw.rds")
saveRDS(single_sensitivity_metrics, "results/simulation/single_sensitivity_metrics.rds")
```

### Leaf-level Results — Single Measure Sensitivity

```{r raw_single_sensitivity_table}
display_raw_results(
  single_sensitivity_raw,
  caption = "Single Measure Sensitivity — per-leaf rejection decisions by signal weight (\u2713 = true signal pair)"
)
```

## Repeated Measures — Run All Methods Across Signal Weights

```{r sensitivity_repeated_benchmark}
repeated_sensitivity_results <- purrr::imap(
  repeated_sensitivity_dfs,
  function(df, sw_name) {
    sw <- as.numeric(sub("sw_", "", sw_name))
    message(sprintf("Benchmarking repeated measures: %s", sw_name))

    spe         <- make_spe_from_df(df)
    spicy_sub   <- repeated_sensitivity_spicy[[sw_name]]$sub
    spicy_major <- repeated_sensitivity_spicy[[sw_name]]$major

    out <- run_all_methods(spicy_sub, spicy_major, spe)
    list(
      raw     = out$raw     %>% mutate(signal_weight = sw, scenario = "Repeated measures"),
      metrics = out$metrics %>% mutate(signal_weight = sw, scenario = "Repeated measures")
    )
  }
)

repeated_sensitivity_raw     <- purrr::map_dfr(repeated_sensitivity_results, "raw")
repeated_sensitivity_metrics <- purrr::map_dfr(repeated_sensitivity_results, "metrics")

saveRDS(repeated_sensitivity_raw,     "results/simulation/repeated_sensitivity_raw.rds")
saveRDS(repeated_sensitivity_metrics, "results/simulation/repeated_sensitivity_metrics.rds")
```

### Leaf-level Results — Repeated Measures Sensitivity

```{r raw_repeated_sensitivity_table}
display_raw_results(
  repeated_sensitivity_raw,
  caption = "Repeated Measures Sensitivity — per-leaf rejection decisions by signal weight (\u2713 = true signal pair)"
)
```

## Sensitivity Analysis — Metric Plots

```{r sensitivity_plots, fig.width=10, fig.height=6}
sensitivity_metrics <- bind_rows(
  single_sensitivity_metrics,
  repeated_sensitivity_metrics
)

METHOD_COLOURS <- c(
  "Flat BH"        = "#E41A1C",
  "Flat BY"        = "#FB9A99",
  "TreeBH"         = "#377EB8",
  "DART"           = "#4DAF4A",
  "pfilter"        = "#984EA3",
  "Ours"           = "#FF7F00",
  "ClusterTrim_BY" = "#A65628"
)

plot_metric <- function(data, metric, y_label,
                        hline = NULL, hline_colour = "grey40") {
  p <- data %>%
    ggplot(aes(x = signal_weight, y = .data[[metric]],
               colour = method, group = method)) +
    geom_line(linewidth = 0.8) +
    geom_point(size = 2.5) +
    facet_wrap(~scenario) +
    scale_colour_manual(values = METHOD_COLOURS, name = "Method") +
    scale_x_continuous(breaks = SIGNAL_WEIGHTS) +
    labs(x = "Signal weight", y = y_label) +
    theme_bw(base_size = 12) +
    theme(legend.position = "bottom")

  if (!is.null(hline))
    p <- p + geom_hline(yintercept = hline, linetype = "dashed",
                        colour = hline_colour)
  p
}

plot_metric(sensitivity_metrics, "TPR", "Power (TPR)") +
  ggtitle("Power across signal weights")

plot_metric(sensitivity_metrics, "FDR", "FDR",
            hline = 0.05, hline_colour = "red") +
  ggtitle("FDR across signal weights")

plot_metric(sensitivity_metrics, "FPR", "FPR") +
  ggtitle("FPR across signal weights")

plot_metric(sensitivity_metrics, "Precision", "Precision") +
  ggtitle("Precision across signal weights")

plot_metric(sensitivity_metrics, "F1", "F1 score") +
  ggtitle("F1 score across signal weights")
```

## Sensitivity Analysis — Metric Tables

```{r sensitivity_table_long}
sensitivity_metrics %>%
  select(scenario, signal_weight, method, TPR, FDR, FPR, Precision, F1) %>%
  arrange(scenario, signal_weight, desc(TPR)) %>%
  knitr::kable(digits = 3,
               caption = "Sensitivity Analysis — Metrics by Signal Weight")
```

```{r sensitivity_table_wide}
sensitivity_metrics %>%
  select(scenario, signal_weight, method, TPR, FDR) %>%
  pivot_wider(
    names_from  = method,
    values_from = c(TPR, FDR),
    names_glue  = "{method}_{.value}"
  ) %>%
  arrange(scenario, signal_weight) %>%
  knitr::kable(digits = 3,
               caption = "Sensitivity Analysis — TPR & FDR Wide Format")
```

# Session Info

```{r session_info}
sessionInfo()
```
