---
title: "simulation_and_benchmarking"
output: html_document
date: "2026-03-29"
---

```{r setup, include=FALSE}
library(tidyverse)
library(SpatialExperiment)
library(spatstat.geom)
library(spatstat.random)
library(spatstat.explore)
library(spicyR)
library(TreeBH)
library(DART)
library(purrr)
library(dplyr)
library(furrr)
source("R/globals.R")
source("R/simulation_helpers.R")
source("R/benchmarking_helpers.R")
source("R/pfilter.R")

# Set up parallel workers — adjust n_workers to however many cores you have
n_workers = 8
plan(multisession, workers = n_workers)
```

# Single Measure Repeated Simulation

```{r}
run_single_measure_replicates = function(
    n_reps        = 20,
    n_subjects    = 50,
    signal_weight = 0.99,
    total_cells   = 5000,
    win           = WIN,
    true_pairs    = TRUE_PAIRS,
    seed          = 42
) {
  set.seed(seed)

  purrr::map_dfr(seq_len(n_reps), function(rep_id) {
    message(sprintf("Running replicate %d / %d", rep_id, n_reps))

    df = purrr::map_dfr(seq_len(n_subjects), function(s) {
      controlPart = simulate_one_subject_consistent(
        subject_id  = s,
        n_images    = 1,
        win         = win,
        condition   = "control",
        add_signal  = FALSE,
        total_cells = total_cells
      )
      signalPart = simulate_one_subject_consistent(
        subject_id    = s + n_subjects,
        n_images      = 1,
        win           = win,
        condition     = "signal",
        add_signal    = TRUE,
        signal_weight = signal_weight,
        total_cells   = total_cells
      )
      bind_rows(controlPart, signalPart)
    })

    spe        = make_spe_from_df(df)
    spicySub     = run_spicy_contrast(spe, "control", "signal", "sub")
    spicyMajor   = run_spicy_contrast(spe, "control", "signal", "major")
    spicyLineage = run_spicy_contrast(spe, "control", "signal", "lineage")

    out = tryCatch(
      run_all_methods(spicySub, spicyMajor, spicyLineage, spe, true_pairs),
      error = function(e) {
        message(sprintf("Replicate %d failed: %s", rep_id, conditionMessage(e)))
        NULL
      }
    )

    if (is.null(out)) return(NULL)

    out$metrics %>%
      mutate(
        rep_id        = rep_id,
        scenario      = "Single measure",
        signal_weight = signal_weight,
        n_subjects    = n_subjects,
        n_images      = 1
      )
  })
}
```

## Running Single Measure Repeated Simulation

```{r}
# singleReplicateMetrics = run_single_measure_replicates(
#   n_reps        = 20,
#   n_subjects    = 50,
#   signal_weight = 0.99
# )
# saveRDS(singleReplicateMetrics, "results/simulation/single_replicate_metrics.rds")
```

## Creating Metrics for the Single Measure Simulation

```{r}
# singleReplicateSummary = singleReplicateMetrics %>%
#   group_by(method) %>%
#   summarise(
#     mean_TPR = mean(TPR, na.rm = TRUE),
#     sd_TPR   = sd(TPR,   na.rm = TRUE),
#     mean_FDR = mean(FDR, na.rm = TRUE),
#     sd_FDR   = sd(FDR,   na.rm = TRUE),
#     mean_F1  = mean(F1,  na.rm = TRUE),
#     sd_F1    = sd(F1,    na.rm = TRUE),
#     n_reps   = n(),
#     .groups  = "drop"
#   ) %>%
#   arrange(desc(mean_TPR))
```

---

# Repeated Measure Simulation

```{r}
run_repeated_measure_replicates = function(
    n_reps        = 20,
    n_subjects    = 50,
    n_images      = 5,
    signal_weight = 0.99,
    total_cells   = 5000,
    win           = WIN,
    true_pairs    = TRUE_PAIRS,
    seed          = 42
) {
  set.seed(seed)

  purrr::map_dfr(seq_len(n_reps), function(rep_id) {
    message(sprintf("Running replicate %d / %d", rep_id, n_reps))

    df = purrr::map_dfr(seq_len(n_subjects), function(s) {
      controlPart = simulate_one_subject_consistent(
        subject_id  = s,
        n_images    = n_images,
        win         = win,
        condition   = "control",
        add_signal  = FALSE,
        total_cells = total_cells
      )
      signalPart = simulate_one_subject_consistent(
        subject_id    = s + n_subjects,
        n_images      = n_images,
        win           = win,
        condition     = "signal",
        add_signal    = TRUE,
        signal_weight = signal_weight,
        total_cells   = total_cells
      )
      bind_rows(controlPart, signalPart)
    })

    spe        = make_spe_from_df(df)
    spicySub     = run_spicy_contrast(spe, "control", "signal", "sub")
    spicyMajor   = run_spicy_contrast(spe, "control", "signal", "major")
    spicyLineage = run_spicy_contrast(spe, "control", "signal", "lineage")

    out = tryCatch(
      run_all_methods(spicySub, spicyMajor, spicyLineage, spe, true_pairs),
      error = function(e) {
        message(sprintf("Replicate %d failed: %s", rep_id, conditionMessage(e)))
        NULL
      }
    )

    if (is.null(out)) return(NULL)

    out$metrics %>%
      mutate(
        rep_id        = rep_id,
        scenario      = "Repeated measure",
        signal_weight = signal_weight,
        n_subjects    = n_subjects,
        n_images      = n_images
      )
  })
}
```

## Running Repeated Measure Simulation

```{r}
# repeatedReplicateMetrics = run_repeated_measure_replicates(
#   n_reps        = 20,
#   n_subjects    = 50,
#   n_images      = 5,
#   signal_weight = 0.99
# )
# saveRDS(repeatedReplicateMetrics, "results/simulation/repeated_replicate_metrics.rds")
```

## Creating Metrics for the Repeated Measure Simulation

```{r}
# repeatedReplicateSummary = repeatedReplicateMetrics %>%
#   group_by(method) %>%
#   summarise(
#     mean_TPR = mean(TPR, na.rm = TRUE),
#     sd_TPR   = sd(TPR,   na.rm = TRUE),
#     mean_FDR = mean(FDR, na.rm = TRUE),
#     sd_FDR   = sd(FDR,   na.rm = TRUE),
#     mean_F1  = mean(F1,  na.rm = TRUE),
#     sd_F1    = sd(F1,    na.rm = TRUE),
#     n_reps   = n(),
#     .groups  = "drop"
#   ) %>%
#   arrange(desc(mean_TPR))
```

---

# Sensitivity Analysis

## Single Measure Sensitivity

```{r}
run_sensitivity_single = function(
    n_reps         = 20,
    n_subjects     = 50,
    signal_weights = c(0.30, 0.50, 0.70, 0.90, 0.99),
    total_cells    = 10000,
    win            = WIN,
    true_pairs     = TRUE_PAIRS,
    cache_dir      = "results/simulation/cache/single"
) {
  dir.create(cache_dir, recursive = TRUE, showWarnings = FALSE)
  dir.create(file.path(cache_dir, "spe"),     recursive = TRUE, showWarnings = FALSE)
  dir.create(file.path(cache_dir, "spicy"),   recursive = TRUE, showWarnings = FALSE)
  dir.create(file.path(cache_dir, "metrics"), recursive = TRUE, showWarnings = FALSE)

  purrr::map_dfr(signal_weights, function(w) {
    message(sprintf("Running signal weight %.2f", w))
    furrr::future_map_dfr(seq_len(n_reps), function(rep_id) {
      wd = getwd()
      source(file.path(wd, "R/globals.R"))
      source(file.path(wd, "R/simulation_helpers.R"))
      source(file.path(wd, "R/benchmarking_helpers.R"))
      source(file.path(wd, "R/pfilter.R"))

      stub        = sprintf("w%.2f_rep%03d", w, rep_id)
      speFile     = file.path(cache_dir, "spe",     paste0(stub, ".rds"))
      spicyFile   = file.path(cache_dir, "spicy",   paste0(stub, ".rds"))
      metricsFile = file.path(cache_dir, "metrics", paste0(stub, ".rds"))

      # ── metrics already done ──────────────────────────────────────────
      if (file.exists(metricsFile)) {
        message(sprintf("  Replicate %d (weight %.2f) metrics loaded from cache", rep_id, w))
        return(readRDS(metricsFile))
      }

      # ── simulate data (or load) ───────────────────────────────────────
      if (file.exists(speFile)) {
        message(sprintf("  Replicate %d (weight %.2f) SPE loaded from cache", rep_id, w))
        spe = readRDS(speFile)
      } else {
        message(sprintf("  Replicate %d / %d — simulating", rep_id, n_reps))
        df = purrr::map_dfr(seq_len(n_subjects), function(s) {
          controlPart = simulate_one_subject_consistent(
            subject_id  = s,
            n_images    = 1,
            win         = win,
            condition   = "control",
            add_signal  = FALSE,
            total_cells = total_cells
          )
          signalPart = simulate_one_subject_consistent(
            subject_id    = s + n_subjects,
            n_images      = 1,
            win           = win,
            condition     = "signal",
            add_signal    = TRUE,
            signal_weight = w,
            total_cells   = total_cells
          )
          bind_rows(controlPart, signalPart)
        })
        spe = make_spe_from_df(df)
        saveRDS(spe, speFile)
        rm(df)
      }

      # ── spicyR (or load) ──────────────────────────────────────────────
      if (file.exists(spicyFile)) {
        message(sprintf("  Replicate %d (weight %.2f) spicy loaded from cache", rep_id, w))
        spicyCache   = readRDS(spicyFile)
        spicySub     = spicyCache$spicySub
        spicyMajor   = spicyCache$spicyMajor
        spicyLineage = spicyCache$spicyLineage
      } else {
        message(sprintf("  Replicate %d / %d — running spicyR", rep_id, n_reps))
        spicySub     = run_spicy_contrast(spe, "control", "signal", "sub",     cores = 8)
        spicyMajor   = run_spicy_contrast(spe, "control", "signal", "major",   cores = 8)
        spicyLineage = run_spicy_contrast(spe, "control", "signal", "lineage", cores = 8)
        saveRDS(
          list(spicySub = spicySub, spicyMajor = spicyMajor,
               spicyLineage = spicyLineage),
          spicyFile
        )
      }

      # ── run methods ───────────────────────────────────────────────────
      result = tryCatch(
        run_all_methods(spicySub, spicyMajor, spicyLineage, spe, true_pairs),
        error = function(e) {
          message(sprintf("  Replicate %d (weight %.2f) failed: %s",
                          rep_id, w, conditionMessage(e)))
          NULL
        }
      )
      if (is.null(result)) return(NULL)

      out = result$metrics %>%
        mutate(
          rep_id        = rep_id,
          scenario      = "Sensitivity - Single measure",
          signal_weight = w,
          n_subjects    = n_subjects,
          n_images      = 1
        )
      saveRDS(out, metricsFile)
      rm(spe, spicySub, spicyMajor, spicyLineage, result)
      gc()
      out
    }, .options = furrr_options(seed = TRUE))
  })
}
```

### Running Single Measure Sensitivity

```{r}
sensitivitySingleMetrics = run_sensitivity_single(
  n_reps         = 20,
  n_subjects     = 50,
  signal_weights = c(0.30, 0.50, 0.70, 0.90, 0.99)
)
saveRDS(sensitivitySingleMetrics, "results/simulation/sensitivity_single_metrics.rds")
```

### Creating Metrics for the Single Measure Sensitivity

```{r}
sensitivitySingleSummary = sensitivitySingleMetrics %>%
  group_by(method, signal_weight) %>%
  summarise(
    mean_TPR = mean(TPR, na.rm = TRUE),
    sd_TPR   = sd(TPR,   na.rm = TRUE),
    mean_FDR = mean(FDR, na.rm = TRUE),
    sd_FDR   = sd(FDR,   na.rm = TRUE),
    mean_F1  = mean(F1,  na.rm = TRUE),
    sd_F1    = sd(F1,    na.rm = TRUE),
    n_reps   = n(),
    .groups  = "drop"
  ) %>%
  arrange(signal_weight, desc(mean_TPR))
```

---

## Repeated Measure Sensitivity

```{r}
run_sensitivity_repeated = function(
    n_reps         = 20,
    n_subjects     = 50,
    n_images       = 5,
    signal_weights = c(0.30, 0.50, 0.70, 0.90, 0.99),
    total_cells    = 10000,
    win            = WIN,
    true_pairs     = TRUE_PAIRS,
    cache_dir      = "results/simulation/cache/repeated"
) {
  dir.create(cache_dir, recursive = TRUE, showWarnings = FALSE)
  dir.create(file.path(cache_dir, "spe"),     recursive = TRUE, showWarnings = FALSE)
  dir.create(file.path(cache_dir, "spicy"),   recursive = TRUE, showWarnings = FALSE)
  dir.create(file.path(cache_dir, "metrics"), recursive = TRUE, showWarnings = FALSE)

  purrr::map_dfr(signal_weights, function(w) {
    message(sprintf("Running signal weight %.2f", w))
    furrr::future_map_dfr(seq_len(n_reps), function(rep_id) {
      wd = getwd()
      source(file.path(wd, "R/globals.R"))
      source(file.path(wd, "R/simulation_helpers.R"))
      source(file.path(wd, "R/benchmarking_helpers.R"))
      source(file.path(wd, "R/pfilter.R"))

      stub        = sprintf("w%.2f_rep%03d", w, rep_id)
      speFile     = file.path(cache_dir, "spe",     paste0(stub, ".rds"))
      spicyFile   = file.path(cache_dir, "spicy",   paste0(stub, ".rds"))
      metricsFile = file.path(cache_dir, "metrics", paste0(stub, ".rds"))

      # ── metrics already done ──────────────────────────────────────────
      if (file.exists(metricsFile)) {
        message(sprintf("  Replicate %d (weight %.2f) metrics loaded from cache", rep_id, w))
        return(readRDS(metricsFile))
      }

      # ── simulate data (or load) ───────────────────────────────────────
      if (file.exists(speFile)) {
        message(sprintf("  Replicate %d (weight %.2f) SPE loaded from cache", rep_id, w))
        spe = readRDS(speFile)
      } else {
        message(sprintf("  Replicate %d / %d — simulating", rep_id, n_reps))
        df = purrr::map_dfr(seq_len(n_subjects), function(s) {
          controlPart = simulate_one_subject_consistent(
            subject_id  = s,
            n_images    = n_images,
            win         = win,
            condition   = "control",
            add_signal  = FALSE,
            total_cells = total_cells
          )
          signalPart = simulate_one_subject_consistent(
            subject_id    = s + n_subjects,
            n_images      = n_images,
            win           = win,
            condition     = "signal",
            add_signal    = TRUE,
            signal_weight = w,
            total_cells   = total_cells
          )
          bind_rows(controlPart, signalPart)
        })
        spe = make_spe_from_df(df)
        saveRDS(spe, speFile)
        rm(df)
      }

      # ── spicyR (or load) ──────────────────────────────────────────────
      if (file.exists(spicyFile)) {
        message(sprintf("  Replicate %d (weight %.2f) spicy loaded from cache", rep_id, w))
        spicyCache   = readRDS(spicyFile)
        spicySub     = spicyCache$spicySub
        spicyMajor   = spicyCache$spicyMajor
        spicyLineage = spicyCache$spicyLineage
      } else {
        message(sprintf("  Replicate %d / %d — running spicyR", rep_id, n_reps))
        spicySub     = run_spicy_contrast(spe, "control", "signal", "sub",     cores = 8)
        spicyMajor   = run_spicy_contrast(spe, "control", "signal", "major",   cores = 8)
        spicyLineage = run_spicy_contrast(spe, "control", "signal", "lineage", cores = 8)
        saveRDS(
          list(spicySub = spicySub, spicyMajor = spicyMajor,
               spicyLineage = spicyLineage),
          spicyFile
        )
      }

      # ── run methods ───────────────────────────────────────────────────
      result = tryCatch(
        run_all_methods(spicySub, spicyMajor, spicyLineage, spe, true_pairs),
        error = function(e) {
          message(sprintf("  Replicate %d (weight %.2f) failed: %s",
                          rep_id, w, conditionMessage(e)))
          NULL
        }
      )
      if (is.null(result)) return(NULL)

      out = result$metrics %>%
        mutate(
          rep_id        = rep_id,
          scenario      = "Sensitivity - Repeated measure",
          signal_weight = w,
          n_subjects    = n_subjects,
          n_images      = n_images
        )
      saveRDS(out, metricsFile)
      rm(spe, spicySub, spicyMajor, spicyLineage, result)
      gc()
      out
    }, .options = furrr_options(seed = TRUE))
  })
}
```

### Running Repeated Measure Sensitivity

```{r}
sensitivityRepeatedMetrics = run_sensitivity_repeated(
  n_reps         = 20,
  n_subjects     = 50,
  n_images       = 5,
  signal_weights = c(0.30, 0.50, 0.70, 0.90, 0.99)
)
saveRDS(sensitivityRepeatedMetrics, "results/simulation/sensitivity_repeated_metrics.rds")
```

### Creating Metrics for the Repeated Measure Sensitivity

```{r}
sensitivityRepeatedSummary = sensitivityRepeatedMetrics %>%
  group_by(method, signal_weight) %>%
  summarise(
    mean_TPR = mean(TPR, na.rm = TRUE),
    sd_TPR   = sd(TPR,   na.rm = TRUE),
    mean_FDR = mean(FDR, na.rm = TRUE),
    sd_FDR   = sd(FDR,   na.rm = TRUE),
    mean_F1  = mean(F1,  na.rm = TRUE),
    sd_F1    = sd(F1,    na.rm = TRUE),
    n_reps   = n(),
    .groups  = "drop"
  ) %>%
  arrange(signal_weight, desc(mean_TPR))
```


---

# Combined Summary Across All Scenarios

```{r}
# allMetrics = bind_rows(
#   singleReplicateMetrics,
#   repeatedReplicateMetrics,
#   sensitivitySingleMetrics,
#   sensitivityRepeatedMetrics
# )
# 
# allSummary = allMetrics %>%
#   group_by(method, scenario, signal_weight) %>%
#   summarise(
#     mean_TPR = mean(TPR, na.rm = TRUE),
#     sd_TPR   = sd(TPR,   na.rm = TRUE),
#     mean_FDR = mean(FDR, na.rm = TRUE),
#     sd_FDR   = sd(FDR,   na.rm = TRUE),
#     mean_F1  = mean(F1,  na.rm = TRUE),
#     sd_F1    = sd(F1,    na.rm = TRUE),
#     n_reps   = n(),
#     .groups  = "drop"
#   ) %>%
#   arrange(scenario, signal_weight, desc(mean_TPR))
# 
# saveRDS(allSummary, "results/simulation/all_summary.rds")
```

# Test

```{r}
# ── single replicate smoke test ───────────────────────────────────────────────
wd = getwd()
source(file.path(wd, "R/globals.R"))
source(file.path(wd, "R/simulation_helpers.R"))
source(file.path(wd, "R/benchmarking_helpers.R"))
source(file.path(wd, "R/pfilter.R"))

set.seed(42)

df = purrr::map_dfr(seq_len(10), function(s) {
  controlPart = simulate_one_subject_consistent(
    subject_id  = s,
    n_images    = 1,
    win         = WIN,
    condition   = "control",
    add_signal  = FALSE,
    total_cells = 10000
  )
  signalPart = simulate_one_subject_consistent(
    subject_id    = s + 10,
    n_images      = 1,
    win           = WIN,
    condition     = "signal",
    add_signal    = TRUE,
    signal_weight = 0.99,
    total_cells   = 10000
  )
  bind_rows(controlPart, signalPart)
})

spe        = make_spe_from_df(df)
spicySub     = run_spicy_contrast(spe, "control", "signal", "sub")
spicyMajor   = run_spicy_contrast(spe, "control", "signal", "major")
spicyLineage = run_spicy_contrast(spe, "control", "signal", "lineage")

testResult   = run_all_methods(spicySub, spicyMajor, spicyLineage, spe, TRUE_PAIRS)

glimpse(testResult$metrics)
```

