---
title: "Simulation"
output: html_document
date: "2025-12-24"
---

# Libraries

```{r setup, include=FALSE}
library(tidyverse)
library(spatstat.geom)
library(spatstat.random)
library(spatstat.explore)
library(SpatialExperiment)
library(spicyR)
library(TreeBH)
library(DART)
source("pfilter.R")
```

# Globals + Helpers

```{r}
set.seed(1)

# --- Global window (define ONCE)
WIN <- owin(c(0, 1000), c(0, 1000))

# --- subtype -> major map (define ONCE)
MAJOR_MAP <- c(CD4="T", CD8="T", Treg="T", B="B", NK="NK")

# --- canonicalize "B__CD4" and "CD4__B" -> "B__CD4"
canonical_pair <- function(x) {
  ab <- strsplit(x, "__", fixed = TRUE)
  vapply(ab, function(z) paste(sort(z), collapse="__"), character(1))
}

leaf_to_family <- function(leaf) {
  ab <- strsplit(leaf, "__", fixed = TRUE)[[1]]
  fam <- sort(c(MAJOR_MAP[ab[1]], MAJOR_MAP[ab[2]]))
  paste(fam, collapse="__")
}

# --- BH helper (your exact logic, reusable)
bh_reject <- function(p, q = 0.05) {
  p <- as.numeric(p)
  m <- length(p)
  o <- order(p)
  ps <- p[o]
  thr <- (seq_len(m) / m) * q
  k <- max(which(ps <= thr), 0)
  out <- rep(FALSE, m)
  if (k > 0) out[o[seq_len(k)]] <- TRUE
  out
}

# --- spatstat intensity normalization
normalize_im <- function(im) im / integral.im(im)

# --- Poisson PP helper
simulate_pp <- function(n, win) {
  pts <- rpoispp(n / area.owin(win), win = win)
  tibble(x = pts$x, y = pts$y)
}

make_ids <- function(i, condition) {
  list(
    imageID   = factor(paste0(condition, "_img_", i)),
    subject   = factor(i),
    condition = factor(condition)
  )
}

finalise_cellData <- function(cellData, ids) {
  cellData %>%
    mutate(
      imageID   = ids$imageID,
      subject   = ids$subject,
      condition = ids$condition,
      cellType_sub = factor(cellType_sub, levels = c("CD4","CD8","Treg","B","NK")),
      cellType_major = factor(cellType_major, levels = c("T","B","NK"))
    )
}

safe_bw_diggle <- function(pp, hmax = 2000) {
  bw.diggle(pp)
}

relabel_condition <- function(out, condition_name, i) {
  out$condition <- factor(condition_name)
  out$imageID   <- factor(paste0(condition_name, "_img_", i))
  out
}

# -----------------------------
# Shared helpers you call later
# -----------------------------
allocate_T_subtypes <- function(n_T) {
  t_sub_props <- c(CD4 = 0.70, CD8 = 0.20, Treg = 0.10)
  n_sub <- as.vector(rmultinom(1, n_T, prob = t_sub_props))
  names(n_sub) <- names(t_sub_props)
  n_sub
}

simulate_T_baseline <- function(n_T, win = WIN) {
  t_pts <- simulate_pp(n_T, win)
  pp_T  <- ppp(t_pts$x, t_pts$y, window = win)
  kT    <- density.ppp(pp_T, sigma = safe_bw_diggle(pp_T))
  list(t_pts = t_pts, kT = kT)
}

# -----------------------------
# Extract leaf-level vectors from spicy objects (aligned + canonicalized)
# -----------------------------
extract_leaf_stats <- function(spicy_sub, conditionB, eps = 1e-12) {
  colB <- paste0("condition", conditionB)

  stopifnot(colB %in% colnames(spicy_sub$p.value))
  stopifnot(colB %in% colnames(spicy_sub$coefficient))

  p_raw <- spicy_sub$p.value[, colB]
  c_raw <- spicy_sub$coefficient[, colB]
  inter <- rownames(spicy_sub$p.value)

  df <- tibble(
    interaction = inter,
    leaf = canonical_pair(interaction),
    pval = as.numeric(p_raw),
    coef = as.numeric(c_raw)
  ) %>%
    group_by(leaf) %>%
    slice(1) %>%          # drops NK__CD4 vs CD4__NK duplicates
    ungroup() %>%
    mutate(pval = pmin(pmax(pval, eps), 1 - eps))

  df
}

extract_major_pvals <- function(spicy_major, conditionB) {
  colB <- paste0("condition", conditionB)

  stopifnot(colB %in% colnames(spicy_major$p.value))

  inter <- rownames(spicy_major$p.value)

  tibble(
    interaction = inter,
    major = canonical_pair(inter),
    p_major = as.numeric(spicy_major$p.value[, colB])
  ) %>%
    group_by(major) %>%
    slice(1) %>%
    ungroup()
}

```

# Simulation Functions

## T / CD4 Attracted to NK

### Leaf-level (CD4 is attracted to NK cells)

- NK is generated first, and kNK is estimated from NK points

- baseline T intensity kT is estimated

- CD8 and Treg are sampled from kT (baseline)

- CD4 is sampled from mixture between kT and kNK

```{r}
simulate_leaf_T_to_NK <- function(i, win = WIN, total_cells = 500, w_nk = 0.70) {
  # set.seed(i)

  props <- c(T=0.75, B=0.15, NK=0.10)
  n_T  <- rpois(1, total_cells * props["T"])
  n_B  <- rpois(1, total_cells * props["B"])
  n_NK <- rpois(1, total_cells * props["NK"])
  n_sub <- allocate_T_subtypes(n_T)

  # NK baseline intensity
  nk_pts <- simulate_pp(n_NK, win)
  pp_NK  <- ppp(nk_pts$x, nk_pts$y, window = win)
  kNK    <- density.ppp(pp_NK, sigma = safe_bw_diggle(pp_NK))

  # T baseline intensity
  Tb <- simulate_T_baseline(n_T, win)
  kT <- Tb$kT

  # CD8/Treg from baseline T
  sim_T <- function(n, label) {
    if (n <= 0) return(NULL)
    pp <- rpoint(n, f = kT, win = win)
    tibble(x=pp$x, y=pp$y, cellType_sub=label, cellType_major="T")
  }
  cd8  <- sim_T(n_sub["CD8"],  "CD8")
  treg <- sim_T(n_sub["Treg"], "Treg")

  # CD4 attracted to NK: mix kernel (T + NK), sample CD4 from it
  k_cd4 <- (1 - w_nk) * normalize_im(kT) + w_nk * normalize_im(kNK)
  pp_cd4 <- rpoint(n_sub["CD4"], f = k_cd4, win = win)
  cd4 <- tibble(x=pp_cd4$x, y=pp_cd4$y, cellType_sub="CD4", cellType_major="T")

  b <- simulate_pp(n_B, win) %>% transmute(x,y, cellType_sub="B", cellType_major="B")
  nk <- nk_pts %>% transmute(x,y, cellType_sub="NK", cellType_major="NK")

  cellData <- bind_rows(cd4, cd8, treg, b, nk)
  finalise_cellData(cellData, make_ids(i, "LeafSignal_TtoNK"))
}
```

## Internal-level (T-cells moves towards NK)

- NK is generated first, and kNK is estimated from NK points

- baseline T intensity kT is estimated

- CD4, CD8, and Treg are sampled from a mixture of kT (baseline) and kNK


```{r}
simulate_internal_T_to_NK <- function(i, win = WIN, total_cells = 500, w_nk = 0.70) {
  # set.seed(i)

  props <- c(T=0.75, B=0.15, NK=0.10)
  n_T  <- rpois(1, total_cells * props["T"])
  n_B  <- rpois(1, total_cells * props["B"])
  n_NK <- rpois(1, total_cells * props["NK"])
  n_sub <- allocate_T_subtypes(n_T)

  # NK baseline intensity
  nk_pts <- simulate_pp(n_NK, win)
  pp_NK  <- ppp(nk_pts$x, nk_pts$y, window = win)
  kNK    <- density.ppp(pp_NK, sigma = safe_bw_diggle(pp_NK))

  # T baseline intensity
  Tb <- simulate_T_baseline(n_T, win)
  kT <- Tb$kT

  # All T subtypes attracted to NK
  k_mix <- (1 - w_nk) * normalize_im(kT) + w_nk * normalize_im(kNK)
  sim_mix <- function(n, label) {
    if (n <= 0) return(NULL)
    pp <- rpoint(n, f = k_mix, win = win)
    tibble(x=pp$x, y=pp$y, cellType_sub=label, cellType_major="T")
  }
  t_sub <- bind_rows(
    sim_mix(n_sub["CD4"],  "CD4"),
    sim_mix(n_sub["CD8"],  "CD8"),
    sim_mix(n_sub["Treg"], "Treg")
  )

  b <- simulate_pp(n_B, win) %>% transmute(x,y, cellType_sub="B", cellType_major="B")
  nk <- nk_pts %>% transmute(x,y, cellType_sub="NK", cellType_major="NK")

  cellData <- bind_rows(t_sub, b, nk)
  finalise_cellData(cellData, make_ids(i, "InternalSignal_TtoNK"))
}
```

## NK attracted to CD4 / T cells

### Leaf-Level (NK moves towards CD4)

- T subtypes are generated from baseline T intensity kT (including CD4)

- from the simulated CD4 points, you estimate kCD4

- NK baseline kNK is estimated from random NK points

- NK is sampled from mixture of kNK and kCD4

```{r}
simulate_leaf_NK_to_CD4 <- function(i, win = WIN, total_cells = 500, w_cd4 = 0.70) {
  # set.seed(i)

  props <- c(T=0.75, B=0.15, NK=0.10)
  n_T  <- rpois(1, total_cells * props["T"])
  n_B  <- rpois(1, total_cells * props["B"])
  n_NK <- rpois(1, total_cells * props["NK"])
  n_sub <- allocate_T_subtypes(n_T)

  # T baseline intensity
  Tb <- simulate_T_baseline(n_T, win)
  kT <- Tb$kT

  # Simulate all T subtypes from baseline T (including CD4)
  sim_T <- function(n, label) {
    if (n <= 0) return(NULL)
    pp <- rpoint(n, f = kT, win = win)
    tibble(x=pp$x, y=pp$y, cellType_sub=label, cellType_major="T")
  }
  cd4  <- sim_T(n_sub["CD4"],  "CD4")
  cd8  <- sim_T(n_sub["CD8"],  "CD8")
  treg <- sim_T(n_sub["Treg"], "Treg")

  # CD4 density (attractor target for NK)
  pp_CD4 <- ppp(cd4$x, cd4$y, window = win)
  kCD4 <- density.ppp(pp_CD4, sigma = safe_bw_diggle(pp_CD4))

  # NK baseline density (for mixing)
  nk_base <- simulate_pp(n_NK, win)
  pp_NK   <- ppp(nk_base$x, nk_base$y, window = win)
  kNK     <- density.ppp(pp_NK, sigma = safe_bw_diggle(pp_NK))

  # NK attracted to CD4
  k_nk <- (1 - w_cd4) * normalize_im(kNK) + w_cd4 * normalize_im(kCD4)
  pp_nk <- rpoint(n_NK, f = k_nk, win = win)
  nk <- tibble(x=pp_nk$x, y=pp_nk$y, cellType_sub="NK", cellType_major="NK")

  b <- simulate_pp(n_B, win) %>% transmute(x,y, cellType_sub="B", cellType_major="B")

  cellData <- bind_rows(cd4, cd8, treg, b, nk)
  finalise_cellData(cellData, make_ids(i, "LeafSignal_NKtoCD4"))
}
```

### Internal (NK Moves towards T Cells)

- T subtypes are generated from baseline T intensity kT (including CD4)

- from the simulated CD4 points, you estimate kCD4

- NK baseline kNK is estimated from random NK points

- NK is sampled from mixture of kNK and kT

```{r}
simulate_internal_NK_to_T <- function(i, win = WIN, total_cells = 500, w_t = 0.70) {
  # set.seed(i)

  props <- c(T=0.75, B=0.15, NK=0.10)
  n_T  <- rpois(1, total_cells * props["T"])
  n_B  <- rpois(1, total_cells * props["B"])
  n_NK <- rpois(1, total_cells * props["NK"])
  n_sub <- allocate_T_subtypes(n_T)

  # T baseline intensity
  Tb <- simulate_T_baseline(n_T, win)
  kT <- Tb$kT

  # Simulate all T subtypes from baseline T
  sim_T <- function(n, label) {
    if (n <= 0) return(NULL)
    pp <- rpoint(n, f = kT, win = win)
    tibble(x=pp$x, y=pp$y, cellType_sub=label, cellType_major="T")
  }
  t_sub <- bind_rows(
    sim_T(n_sub["CD4"],  "CD4"),
    sim_T(n_sub["CD8"],  "CD8"),
    sim_T(n_sub["Treg"], "Treg")
  )

  # NK baseline density (for mixing)
  nk_base <- simulate_pp(n_NK, win)
  pp_NK   <- ppp(nk_base$x, nk_base$y, window = win)
  kNK     <- density.ppp(pp_NK, sigma = safe_bw_diggle(pp_NK))

  # NK attracted to overall T intensity
  k_nk <- (1 - w_t) * normalize_im(kNK) + w_t * normalize_im(kT)
  pp_nk <- rpoint(n_NK, f = k_nk, win = win)
  nk <- tibble(x=pp_nk$x, y=pp_nk$y, cellType_sub="NK", cellType_major="NK")

  b <- simulate_pp(n_B, win) %>% transmute(x,y, cellType_sub="B", cellType_major="B")

  cellData <- bind_rows(t_sub, b, nk)
  finalise_cellData(cellData, make_ids(i, "InternalSignal_NKtoT"))
}
```

## Simulate Null

- T subtypes all from baseline kT

- B and NK are uniform Poisson

- No mixing was seen

```{r}
simulate_null <- function(i, win = WIN, total_cells = 500) {
  # set.seed(i)

  props <- c(T=0.75, B=0.15, NK=0.10)
  n_T  <- rpois(1, total_cells * props["T"])
  n_B  <- rpois(1, total_cells * props["B"])
  n_NK <- rpois(1, total_cells * props["NK"])
  n_sub <- allocate_T_subtypes(n_T)

  # baseline T intensity
  Tb <- simulate_T_baseline(n_T, win)
  kT <- Tb$kT

  sim_T <- function(n, label) {
    if (n <= 0) return(NULL)
    pp <- rpoint(n, f = kT, win = win)
    tibble(x=pp$x, y=pp$y, cellType_sub=label, cellType_major="T")
  }

  t_sub <- bind_rows(
    sim_T(n_sub["CD4"],  "CD4"),
    sim_T(n_sub["CD8"],  "CD8"),
    sim_T(n_sub["Treg"], "Treg")
  )

  b  <- simulate_pp(n_B,  win) %>% transmute(x,y, cellType_sub="B",  cellType_major="B")
  nk <- simulate_pp(n_NK, win) %>% transmute(x,y, cellType_sub="NK", cellType_major="NK")

  cellData <- bind_rows(t_sub, b, nk)
  finalise_cellData(cellData, make_ids(i, "Null"))
}
```


# Generate the Dataset

## Creating Wrappers for different spatial strength

This is where we have null vs diferent spatial signal strength

### T / CD4 -> NK

#### CD4 -> NK

```{r}
simulate_leaf_T_to_NK_weak <- function(i, win = WIN, total_cells = 500) {
  out <- simulate_leaf_T_to_NK(i, win = win, total_cells = total_cells, w_nk = 0.30)
  out$condition <- factor("LeafSignal_TtoNK_weak")
  out$imageID   <- factor(paste0("LeafSignal_TtoNK_weak_img_", i))
  out
}

simulate_leaf_T_to_NK_medium <- function(i, win = WIN, total_cells = 500) {
  out <- simulate_leaf_T_to_NK(i, win = win, total_cells = total_cells, w_nk = 0.50)
  out$condition <- factor("LeafSignal_TtoNK_medium")
  out$imageID   <- factor(paste0("LeafSignal_TtoNK_medium_img_", i))
  out
}

simulate_leaf_T_to_NK_strong <- function(i, win = WIN, total_cells = 500) {
  out <- simulate_leaf_T_to_NK(i, win = win, total_cells = total_cells, w_nk = 0.80)
  out$condition <- factor("LeafSignal_TtoNK_strong")
  out$imageID   <- factor(paste0("LeafSignal_TtoNK_strong_img_", i))
  out
}
```

#### T -> NK

```{r}
simulate_internal_T_to_NK_weak <- function(i, win = WIN, total_cells = 500) {
  out <- simulate_internal_T_to_NK(i, win = win, total_cells = total_cells, w_nk = 0.30)
  out$condition <- factor("InternalSignal_TtoNK_weak")
  out$imageID   <- factor(paste0("InternalSignal_TtoNK_weak_img_", i))
  out
}

simulate_internal_T_to_NK_medium <- function(i, win = WIN, total_cells = 500) {
  out <- simulate_internal_T_to_NK(i, win = win, total_cells = total_cells, w_nk = 0.50)
  out$condition <- factor("InternalSignal_TtoNK_medium")
  out$imageID   <- factor(paste0("InternalSignal_TtoNK_medium_img_", i))
  out
}

simulate_internal_T_to_NK_strong <- function(i, win = WIN, total_cells = 500) {
  out <- simulate_internal_T_to_NK(i, win = win, total_cells = total_cells, w_nk = 0.80)
  out$condition <- factor("InternalSignal_TtoNK_strong")
  out$imageID   <- factor(paste0("InternalSignal_TtoNK_strong_img_", i))
  out
}
```

### NK -> CD4 / T

#### NK -> CD4

```{r}
simulate_leaf_NK_to_CD4_weak <- function(i, win = WIN, total_cells = 500) {
  out <- simulate_leaf_NK_to_CD4(i, win = win, total_cells = total_cells, w_cd4 = 0.30)
  out$condition <- factor("LeafSignal_NKtoCD4_weak")
  out$imageID   <- factor(paste0("LeafSignal_NKtoCD4_weak_img_", i))
  out
}

simulate_leaf_NK_to_CD4_medium <- function(i, win = WIN, total_cells = 500) {
  out <- simulate_leaf_NK_to_CD4(i, win = win, total_cells = total_cells, w_cd4 = 0.50)
  out$condition <- factor("LeafSignal_NKtoCD4_medium")
  out$imageID   <- factor(paste0("LeafSignal_NKtoCD4_medium_img_", i))
  out
}

simulate_leaf_NK_to_CD4_strong <- function(i, win = WIN, total_cells = 500) {
  out <- simulate_leaf_NK_to_CD4(i, win = win, total_cells = total_cells, w_cd4 = 0.80)
  out$condition <- factor("LeafSignal_NKtoCD4_strong")
  out$imageID   <- factor(paste0("LeafSignal_NKtoCD4_strong_img_", i))
  out
}
```

#### NK -> T

```{r}
simulate_internal_NK_to_T_weak <- function(i, win = WIN, total_cells = 500) {
  out <- simulate_internal_NK_to_T(i, win = win, total_cells = total_cells, w_t = 0.30)
  out$condition <- factor("InternalSignal_NKtoT_weak")
  out$imageID   <- factor(paste0("InternalSignal_NKtoT_weak_img_", i))
  out
}

simulate_internal_NK_to_T_medium <- function(i, win = WIN, total_cells = 500) {
  out <- simulate_internal_NK_to_T(i, win = win, total_cells = total_cells, w_t = 0.50)
  out$condition <- factor("InternalSignal_NKtoT_medium")
  out$imageID   <- factor(paste0("InternalSignal_NKtoT_medium_img_", i))
  out
}

simulate_internal_NK_to_T_strong <- function(i, win = WIN, total_cells = 500) {
  out <- simulate_internal_NK_to_T(i, win = win, total_cells = total_cells, w_t = 0.80)
  out$condition <- factor("InternalSignal_NKtoT_strong")
  out$imageID   <- factor(paste0("InternalSignal_NKtoT_strong_img_", i))
  out
}
```

## Generating Mixed Control and Treatment 

### Leaf T / CD4 -> NK

#### Control

```{r}
simulate_leaf_T_to_NK_control <- function(
  i,
  control_strength = 0.20,
  treatment_strength = 0.50,
  win = WIN,
  total_cells = 500
) {
  out <- simulate_leaf_T_to_NK(
    i,
    win = win,
    total_cells = total_cells,
    w_nk = control_strength
  )

  condition_name <- paste0(
    "Control_Leaf_TtoNK_",
    gsub("\\.", "p", as.character(control_strength))
  )

  relabel_condition(out, condition_name, i)
}
```

#### Treatment

```{r}
simulate_leaf_T_to_NK_treatment <- function(
  i,
  control_strength = 0.20,
  treatment_strength = 0.50,
  win = WIN,
  total_cells = 500
) {
  out <- simulate_leaf_T_to_NK(
    i,
    win = win,
    total_cells = total_cells,
    w_nk = treatment_strength
  )

  condition_name <- paste0(
    "Treatment_Leaf_TtoNK_",
    gsub("\\.", "p", as.character(treatment_strength))
  )

  relabel_condition(out, condition_name, i)
}
```

### Internal T -> NK

#### Control

```{r}
simulate_internal_T_to_NK_control <- function(
  i,
  control_strength = 0.20,
  treatment_strength = 0.50,
  win = WIN,
  total_cells = 500
) {
  out <- simulate_internal_T_to_NK(
    i,
    win = win,
    total_cells = total_cells,
    w_nk = control_strength
  )

  condition_name <- paste0(
    "Control_Internal_TtoNK_",
    gsub("\\.", "p", as.character(control_strength))
  )

  relabel_condition(out, condition_name, i)
}
```

#### Treatment

```{r}
simulate_internal_T_to_NK_treatment <- function(
  i,
  control_strength = 0.20,
  treatment_strength = 0.50,
  win = WIN,
  total_cells = 500
) {
  out <- simulate_internal_T_to_NK(
    i,
    win = win,
    total_cells = total_cells,
    w_nk = treatment_strength
  )

  condition_name <- paste0(
    "Treatment_Internal_TtoNK_",
    gsub("\\.", "p", as.character(treatment_strength))
  )

  relabel_condition(out, condition_name, i)
}
```

### Leaf NK -> CD4

#### Control

```{r}
simulate_leaf_NK_to_CD4_control <- function(
  i,
  control_strength = 0.20,
  treatment_strength = 0.50,
  win = WIN,
  total_cells = 500
) {
  out <- simulate_leaf_NK_to_CD4(
    i,
    win = win,
    total_cells = total_cells,
    w_cd4 = control_strength
  )

  condition_name <- paste0(
    "Control_Leaf_NKtoCD4_",
    gsub("\\.", "p", as.character(control_strength))
  )

  relabel_condition(out, condition_name, i)
}
```

#### Treatment

```{r}
simulate_leaf_NK_to_CD4_treatment <- function(
  i,
  control_strength = 0.20,
  treatment_strength = 0.50,
  win = WIN,
  total_cells = 500
) {
  out <- simulate_leaf_NK_to_CD4(
    i,
    win = win,
    total_cells = total_cells,
    w_cd4 = treatment_strength
  )

  condition_name <- paste0(
    "Treatment_Leaf_NKtoCD4_",
    gsub("\\.", "p", as.character(treatment_strength))
  )

  relabel_condition(out, condition_name, i)
}
```

### Internal NK -> T

#### Control

```{r}
simulate_internal_NK_to_T_control <- function(
  i,
  control_strength = 0.20,
  treatment_strength = 0.50,
  win = WIN,
  total_cells = 500
) {
  out <- simulate_internal_NK_to_T(
    i,
    win = win,
    total_cells = total_cells,
    w_t = control_strength
  )

  condition_name <- paste0(
    "Control_Internal_NKtoT_",
    gsub("\\.", "p", as.character(control_strength))
  )

  relabel_condition(out, condition_name, i)
}
```

#### Treatment

```{r}
simulate_internal_NK_to_T_treatment <- function(
  i,
  control_strength = 0.20,
  treatment_strength = 0.50,
  win = WIN,
  total_cells = 500
) {
  out <- simulate_internal_NK_to_T(
    i,
    win = win,
    total_cells = total_cells,
    w_t = treatment_strength
  )

  condition_name <- paste0(
    "Treatment_Internal_NKtoT_",
    gsub("\\.", "p", as.character(treatment_strength))
  )

  relabel_condition(out, condition_name, i)
}
```

## Running the Simulation

### Running the Null Vs. Various Spatial Strength

```{r}
N_PER_COND <- 100

# cellData_all <- bind_rows(
#   lapply(1:N_PER_COND, simulate_null),
# 
#   # Part A: T/CD4 -> NK
#   lapply(1:N_PER_COND, simulate_leaf_T_to_NK),
#   lapply(1:N_PER_COND, simulate_internal_T_to_NK),
# 
#   # Part B: NK -> CD4/T
#   lapply(1:N_PER_COND, simulate_leaf_NK_to_CD4),
#   lapply(1:N_PER_COND, simulate_internal_NK_to_T)
# )

cellData_all <- bind_rows(
  lapply(1:N_PER_COND, simulate_null),

  # -----------------------------
  # Part A: T/CD4 -> NK
  # -----------------------------
  lapply(1:N_PER_COND, simulate_leaf_T_to_NK_weak),
  lapply(1:N_PER_COND, simulate_leaf_T_to_NK_medium),
  lapply(1:N_PER_COND, simulate_leaf_T_to_NK_strong),

  lapply(1:N_PER_COND, simulate_internal_T_to_NK_weak),
  lapply(1:N_PER_COND, simulate_internal_T_to_NK_medium),
  lapply(1:N_PER_COND, simulate_internal_T_to_NK_strong),

  # -----------------------------
  # Part B: NK -> CD4 / T
  # -----------------------------
  lapply(1:N_PER_COND, simulate_leaf_NK_to_CD4_weak),
  lapply(1:N_PER_COND, simulate_leaf_NK_to_CD4_medium),
  lapply(1:N_PER_COND, simulate_leaf_NK_to_CD4_strong),

  lapply(1:N_PER_COND, simulate_internal_NK_to_T_weak),
  lapply(1:N_PER_COND, simulate_internal_NK_to_T_medium),
  lapply(1:N_PER_COND, simulate_internal_NK_to_T_strong)
)

table(cellData_all$condition)
```

#### Verifying it works

```{r}
# one row per image-condition
img_summary <- cellData_all %>%
  group_by(condition, imageID) %>%
  summarise(
    n_cells = n(),
    n_CD4   = sum(cellType_sub == "CD4"),
    n_CD8   = sum(cellType_sub == "CD8"),
    n_Treg  = sum(cellType_sub == "Treg"),
    n_B     = sum(cellType_sub == "B"),
    n_NK    = sum(cellType_sub == "NK"),
    xmin = min(x), xmax = max(x),
    ymin = min(y), ymax = max(y),
    .groups = "drop"
  )

img_summary %>%
  group_by(condition) %>%
  summarise(
    n_images = n(),
    mean_cells = mean(n_cells),
    mean_CD4 = mean(n_CD4),
    mean_NK  = mean(n_NK)
  )

plot_image <- function(dat, img) {
  dat %>%
    filter(imageID == img) %>%
    ggplot(aes(x, y, color = cellType_sub)) +
    geom_point(size = 1, alpha = 0.8) +
    coord_fixed() +
    theme_bw() +
    ggtitle(as.character(img))
}

# example
some_imgs <- cellData_all %>%
  distinct(condition, imageID) %>%
  group_by(condition) %>%
  dplyr::slice(1) %>%
  ungroup()

plot_image(cellData_all, some_imgs$imageID[1])
```

```{r}
mean_nn_dist <- function(dat, from_type, to_type, major = FALSE) {
  a <- if (major) dat %>% filter(cellType_major == from_type) else dat %>% filter(cellType_sub == from_type)
  b <- if (major) dat %>% filter(cellType_major == to_type) else dat %>% filter(cellType_sub == to_type)

  if (nrow(a) == 0 || nrow(b) == 0) return(NA_real_)

  XA <- cbind(a$x, a$y)
  XB <- cbind(b$x, b$y)

  dmat <- as.matrix(dist(rbind(XA, XB)))
  na <- nrow(XA)
  nb <- nrow(XB)

  dAB <- dmat[1:na, (na + 1):(na + nb), drop = FALSE]
  mean(apply(dAB, 1, min))
}

nn_summary <- cellData_all %>%
  group_by(condition, imageID) %>%
  group_modify(~{
    dat <- .x
    tibble(
      CD4_to_NK = mean_nn_dist(dat, "CD4", "NK"),
      NK_to_CD4 = mean_nn_dist(dat, "NK", "CD4"),
      NK_to_T   = mean_nn_dist(dat, "NK", "T", major = TRUE),
      CD8_to_NK = mean_nn_dist(dat, "CD8", "NK"),
      Treg_to_NK = mean_nn_dist(dat, "Treg", "NK")
    )
  }) %>%
  ungroup()

nn_summary %>%
  group_by(condition) %>%
  summarise(across(c(CD4_to_NK, NK_to_CD4, NK_to_T, CD8_to_NK, Treg_to_NK), mean, na.rm = TRUE))

nn_summary %>%
  filter(str_detect(condition, "LeafSignal_TtoNK")) %>%
  mutate(strength = case_when(
    str_detect(condition, "weak") ~ "weak",
    str_detect(condition, "medium") ~ "medium",
    str_detect(condition, "strong") ~ "strong"
  )) %>%
  group_by(strength) %>%
  summarise(mean_CD4_to_NK = mean(CD4_to_NK, na.rm = TRUE))
```


### Mixed Spatial Strength in Treatment and Control

```{r}
N_PER_COND <- 100

cellData_strength_pairs <- bind_rows(
  lapply(1:N_PER_COND, simulate_leaf_T_to_NK_control,
         control_strength = 0.20, treatment_strength = 0.50),
  lapply(1:N_PER_COND, simulate_leaf_T_to_NK_treatment,
         control_strength = 0.20, treatment_strength = 0.50),

  lapply(1:N_PER_COND, simulate_internal_T_to_NK_control,
         control_strength = 0.20, treatment_strength = 0.50),
  lapply(1:N_PER_COND, simulate_internal_T_to_NK_treatment,
         control_strength = 0.20, treatment_strength = 0.50),

  lapply(1:N_PER_COND, simulate_leaf_NK_to_CD4_control,
         control_strength = 0.20, treatment_strength = 0.50),
  lapply(1:N_PER_COND, simulate_leaf_NK_to_CD4_treatment,
         control_strength = 0.20, treatment_strength = 0.50),

  lapply(1:N_PER_COND, simulate_internal_NK_to_T_control,
         control_strength = 0.20, treatment_strength = 0.50),
  lapply(1:N_PER_COND, simulate_internal_NK_to_T_treatment,
         control_strength = 0.20, treatment_strength = 0.50)
)

table(cellData_strength_pairs$condition)
```

#### Verifying it all works

```{r}
# one row per image-condition
img_summary_pairs <- cellData_strength_pairs %>%
  group_by(condition, imageID) %>%
  summarise(
    n_cells = n(),
    n_CD4   = sum(cellType_sub == "CD4"),
    n_CD8   = sum(cellType_sub == "CD8"),
    n_Treg  = sum(cellType_sub == "Treg"),
    n_B     = sum(cellType_sub == "B"),
    n_NK    = sum(cellType_sub == "NK"),
    xmin = min(x), xmax = max(x),
    ymin = min(y), ymax = max(y),
    .groups = "drop"
  )

# quick sanity check by condition
img_summary_pairs %>%
  group_by(condition) %>%
  summarise(
    n_images = n(),
    mean_cells = mean(n_cells),
    mean_CD4 = mean(n_CD4),
    mean_NK  = mean(n_NK),
    .groups = "drop"
  )

# plotting helper
plot_image <- function(dat, img) {
  dat %>%
    filter(imageID == img) %>%
    ggplot(aes(x, y, color = cellType_sub)) +
    geom_point(size = 1, alpha = 0.8) +
    coord_fixed() +
    theme_bw() +
    ggtitle(as.character(img))
}

# plot one example image from each condition
some_imgs_pairs <- cellData_strength_pairs %>%
  distinct(condition, imageID) %>%
  group_by(condition) %>%
  dplyr::slice(1) %>%
  ungroup()

plot_image(cellData_strength_pairs, some_imgs_pairs$imageID[1])

# nearest-neighbour distance helper
mean_nn_dist <- function(dat, from_type, to_type, major = FALSE) {
  a <- if (major) dat %>% filter(cellType_major == from_type) else dat %>% filter(cellType_sub == from_type)
  b <- if (major) dat %>% filter(cellType_major == to_type) else dat %>% filter(cellType_sub == to_type)

  if (nrow(a) == 0 || nrow(b) == 0) return(NA_real_)

  XA <- cbind(a$x, a$y)
  XB <- cbind(b$x, b$y)

  dmat <- as.matrix(dist(rbind(XA, XB)))
  na <- nrow(XA)
  nb <- nrow(XB)

  dAB <- dmat[1:na, (na + 1):(na + nb), drop = FALSE]
  mean(apply(dAB, 1, min))
}

# per-image NN summaries
nn_summary_pairs <- cellData_strength_pairs %>%
  group_by(condition, imageID) %>%
  group_modify(~{
    dat <- .x
    tibble(
      CD4_to_NK  = mean_nn_dist(dat, "CD4", "NK"),
      NK_to_CD4  = mean_nn_dist(dat, "NK", "CD4"),
      NK_to_T    = mean_nn_dist(dat, "NK", "T", major = TRUE),
      CD8_to_NK  = mean_nn_dist(dat, "CD8", "NK"),
      Treg_to_NK = mean_nn_dist(dat, "Treg", "NK")
    )
  }) %>%
  ungroup()

# average NN distances by condition
nn_summary_pairs %>%
  group_by(condition) %>%
  summarise(
    across(c(CD4_to_NK, NK_to_CD4, NK_to_T, CD8_to_NK, Treg_to_NK), mean, na.rm = TRUE),
    .groups = "drop"
  )
```


```{r}
nn_summary_pairs_labeled <- nn_summary_pairs %>%
  mutate(
    scenario = case_when(
      str_detect(condition, "Leaf_TtoNK")     ~ "Leaf_TtoNK",
      str_detect(condition, "Internal_TtoNK") ~ "Internal_TtoNK",
      str_detect(condition, "Leaf_NKtoCD4")   ~ "Leaf_NKtoCD4",
      str_detect(condition, "Internal_NKtoT") ~ "Internal_NKtoT"
    ),
    group = case_when(
      str_detect(condition, "^Control")   ~ "Control",
      str_detect(condition, "^Treatment") ~ "Treatment"
    ),
    strength = str_extract(condition, "[0-9]p[0-9]+") %>%
      str_replace("p", ".") %>%
      as.numeric()
  )

nn_summary_pairs_labeled %>%
  group_by(scenario, group, strength) %>%
  summarise(
    mean_CD4_to_NK = mean(CD4_to_NK, na.rm = TRUE),
    mean_NK_to_CD4 = mean(NK_to_CD4, na.rm = TRUE),
    mean_NK_to_T   = mean(NK_to_T, na.rm = TRUE),
    .groups = "drop"
  )
```

# Building the SpatialExperiment

```{r}
spe <- SpatialExperiment(
  spatialCoords = as.matrix(cellData_all[, c("x", "y")]),
  colData = cellData_all %>%
    select(imageID, subject, condition, cellType_sub, cellType_major)
)

spe_pairs <- SpatialExperiment(
  spatialCoords = as.matrix(cellData_strength_pairs[, c("x", "y")]),
  colData = cellData_strength_pairs %>%
    select(imageID, subject, condition, cellType_sub, cellType_major)
)
```

```{r}
saveRDS(spe, "spe.rds")
saveRDS(spe_pairs, "spe_pairs.rds")
```


# running spicyR: Null vs each condition

```{r}
run_spicy_contrast <- function(spe, cond_A = "Null", cond_B, level = c("sub", "major")) {
  level <- match.arg(level)
  cell_col <- if (level == "sub") "cellType_sub" else "cellType_major"

  spe2 <- spe[, spe$condition %in% c(cond_A, cond_B)]
  spe2$condition <- droplevels(spe2$condition)
  spe2$imageID   <- droplevels(spe2$imageID)
  spe2[[cell_col]] <- droplevels(spe2[[cell_col]])

  spicy(
    spe2,
    condition = "condition",
    subject   = "subject",
    cellType  = cell_col,
    imageID   = "imageID"
  )
}
```

# Running the Simulation

```{r}
conds <- c(
  "LeafSignal_TtoNK_weak",    "LeafSignal_TtoNK_medium",    "LeafSignal_TtoNK_strong",
  "InternalSignal_TtoNK_weak","InternalSignal_TtoNK_medium","InternalSignal_TtoNK_strong",
  "LeafSignal_NKtoCD4_weak",  "LeafSignal_NKtoCD4_medium",  "LeafSignal_NKtoCD4_strong",
  "InternalSignal_NKtoT_weak","InternalSignal_NKtoT_medium","InternalSignal_NKtoT_strong"
)

spicy_sub_list   <- setNames(lapply(conds, \(cc) run_spicy_contrast(spe, "Null", cc, "sub")),   conds)
spicy_major_list <- setNames(lapply(conds, \(cc) run_spicy_contrast(spe, "Null", cc, "major")), conds)

# Quick sanity: top pairs
lapply(spicy_sub_list, topPairs)
lapply(spicy_major_list, topPairs)
```

```{r}
example_imgs <- cellData_all %>%
  distinct(imageID, condition) %>%
  group_by(condition) %>%
  slice_sample(n = 1) %>%
  pull(imageID)

cellData_all %>%
  filter(imageID %in% example_imgs) %>%
  ggplot(aes(x, y, color = cellType_sub)) +
  geom_point(size = 0.6, alpha = 0.7) +
  coord_fixed() +
  facet_wrap(~ condition) +
  theme_minimal()
```

  