---
title: "Simulation Benchmarking and Sensitivity Analysis"
output: html_document
params:
  n_reps: 20
  n_control: 10
  n_signal: 10
  total_cells: 5000
  signal_weight: 0.70
  q: 0.05
  control_label: "control"
  signal_label: "signal"
  out_dir: "results"
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)

library(tidyverse)
library(SpatialExperiment)
library(S4Vectors)
library(purrr)
library(dplyr)
library(tibble)

source("R/globals.R")
source("R/simulation_helpers.R")
source("R/benchmarking_helpers.R")

# WIN is used by run_one_simulation() but is not defined in globals.R,
# so define it here once.
WIN <- spatstat.geom::owin(c(0, 1000), c(0, 1000))

dir.create(params$out_dir, recursive = TRUE, showWarnings = FALSE)

`%||%` <- function(x, y) if (is.null(x)) y else x
```

# Overview

This notebook does two things:

1. Runs the main simulation and benchmarking workflow `r params$n_reps` times.
2. Runs a sensitivity analysis where each scenario is also repeated `r params$n_reps` times and benchmarked.

The code is written to use your existing helper files, especially:

- `run_one_simulation()` from `simulation_helpers.R`
- the benchmark helper functions from `benchmarking_helpers.R`

---

# Core helper functions

```{r helpers}

```

---

# Main benchmark: 20 repeated simulations

```{r main-benchmark}
main_benchmark <- run_benchmark_scenario(
  scenario = "main",
  n_reps = params$n_reps,
  n_control = params$n_control,
  n_signal = params$n_signal,
  total_cells = params$total_cells,
  signal_weight = params$signal_weight,
  q = params$q
)

main_summary <- main_benchmark$summary
main_detailed <- main_benchmark$detailed

write_csv(main_summary, file.path(params$out_dir, "main_benchmark_summary.csv"))
write_csv(main_detailed, file.path(params$out_dir, "main_benchmark_detailed.csv"))

main_summary
```

## Main benchmark summary by method

```{r main-summary-table}
main_summary %>%
  group_by(method) %>%
  summarise(
    mean_rejections = mean(n_reject),
    sd_rejections   = sd(n_reject),
    mean_power      = mean(power, na.rm = TRUE),
    sd_power        = sd(power, na.rm = TRUE),
    mean_fdp        = mean(fdp),
    sd_fdp          = sd(fdp),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_power), mean_fdp)
```

## Main benchmark plots

```{r main-plots}
main_summary %>%
  ggplot(aes(x = method, y = power)) +
  geom_boxplot() +
  coord_flip() +
  theme_bw() +
  labs(title = "Power across repeated benchmark runs", x = NULL, y = "Power")
```

```{r main-plots-fdp}
main_summary %>%
  ggplot(aes(x = method, y = fdp)) +
  geom_boxplot() +
  coord_flip() +
  theme_bw() +
  labs(title = "FDP across repeated benchmark runs", x = NULL, y = "False discovery proportion")
```

---

# Sensitivity analysis

Edit this tibble to add, remove, or change the scenarios you want to test.
Each row is one scenario, and **each scenario is run `r params$n_reps` times**.

```{r sensitivity-plan}
sensitivity_plan <- tribble(
  ~scenario,                ~n_control, ~n_signal, ~total_cells, ~signal_weight,
  "baseline",                      10,        10,         5000,           0.70,
  "weaker_signal",                 10,        10,         5000,           0.30,
  "stronger_signal",               10,        10,         5000,           0.90,
  "fewer_cells",                   10,        10,         3000,           0.70,
  "more_cells",                    10,        10,         7000,           0.70,
  "smaller_sample_size",            5,         5,         5000,           0.70,
  "larger_sample_size",            15,        15,         5000,           0.70
)

sensitivity_plan
```

```{r run-sensitivity}
sensitivity_results <- pmap(
  sensitivity_plan,
  function(scenario, n_control, n_signal, total_cells, signal_weight) {
    run_benchmark_scenario(
      scenario = scenario,
      n_reps = params$n_reps,
      n_control = n_control,
      n_signal = n_signal,
      total_cells = total_cells,
      signal_weight = signal_weight,
      q = params$q
    )
  }
)

names(sensitivity_results) <- sensitivity_plan$scenario

sensitivity_summary <- bind_rows(map(sensitivity_results, "summary"))
sensitivity_detailed <- bind_rows(map(sensitivity_results, "detailed"))

write_csv(sensitivity_summary, file.path(params$out_dir, "sensitivity_summary.csv"))
write_csv(sensitivity_detailed, file.path(params$out_dir, "sensitivity_detailed.csv"))
```

## Sensitivity summary table

```{r sensitivity-summary-table}
sensitivity_summary %>%
  group_by(scenario, method) %>%
  summarise(
    mean_rejections = mean(n_reject),
    mean_power      = mean(power, na.rm = TRUE),
    mean_fdp        = mean(fdp),
    .groups = "drop"
  ) %>%
  arrange(scenario, desc(mean_power), mean_fdp)
```

## Sensitivity plots

```{r sensitivity-power-plot}
sensitivity_summary %>%
  ggplot(aes(x = scenario, y = power, fill = method)) +
  geom_boxplot(outlier.alpha = 0.4) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Sensitivity analysis: power", x = NULL, y = "Power")
```

```{r sensitivity-fdp-plot}
sensitivity_summary %>%
  ggplot(aes(x = scenario, y = fdp, fill = method)) +
  geom_boxplot(outlier.alpha = 0.4) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Sensitivity analysis: FDP", x = NULL, y = "False discovery proportion")
```

---

# Save a compact comparison table

```{r final-exports}
comparison_table <- sensitivity_summary %>%
  group_by(scenario, method) %>%
  summarise(
    avg_power = mean(power, na.rm = TRUE),
    avg_fdp = mean(fdp),
    avg_rejections = mean(n_reject),
    .groups = "drop"
  )

write_csv(comparison_table, file.path(params$out_dir, "comparison_table.csv"))
comparison_table
```

# Notes

- The simulator currently injects signal by making **NK** cells spatially attracted to **CD4_T**, **CD8_T**, and **Memory_T**, so those three leaf interactions are treated as the ground truth positives for benchmarking.
- Because your helper scripts do not define `WIN`, it is defined near the top of this notebook.
- If you already have a custom `run_spicy_contrast()` helper from your older `.Rmd` files, source it before the benchmarking chunks and this notebook will use it automatically.
- If your local `spicyR::spicy()` argument names differ from the fallback wrapper, just edit the `run_spicy_pair()` helper once and the rest of the notebook can stay the same.
