---
title: "Simulation"
output: html_document
date: "2025-12-24"
---

# Libraries

```{r setup, include=FALSE}
library(tidyverse)
library(spatstat.geom)
library(spatstat.random)
library(spatstat.explore)
library(SpatialExperiment)
library(spicyR)
library(TreeBH)
library(DART)
source("pfilter.R")
```

# Globals + Helpers

```{r}
set.seed(1)

# --- Global window (define ONCE)
WIN <- owin(c(0, 1000), c(0, 1000))

# --- subtype -> major map (define ONCE)
MAJOR_MAP <- c(CD4="T", CD8="T", Treg="T", B="B", NK="NK")

# --- canonicalize "B__CD4" and "CD4__B" -> "B__CD4"
canonical_pair <- function(x) {
  ab <- strsplit(x, "__", fixed = TRUE)
  vapply(ab, function(z) paste(sort(z), collapse="__"), character(1))
}

leaf_to_family <- function(leaf) {
  ab <- strsplit(leaf, "__", fixed = TRUE)[[1]]
  fam <- sort(c(MAJOR_MAP[ab[1]], MAJOR_MAP[ab[2]]))
  paste(fam, collapse="__")
}

# --- BH helper (your exact logic, reusable)
bh_reject <- function(p, q = 0.05) {
  p <- as.numeric(p)
  m <- length(p)
  o <- order(p)
  ps <- p[o]
  thr <- (seq_len(m) / m) * q
  k <- max(which(ps <= thr), 0)
  out <- rep(FALSE, m)
  if (k > 0) out[o[seq_len(k)]] <- TRUE
  out
}

# --- spatstat intensity normalization
normalize_im <- function(im) im / integral.im(im)

# --- Poisson PP helper
simulate_pp <- function(n, win) {
  pts <- rpoispp(n / area.owin(win), win = win)
  tibble(x = pts$x, y = pts$y)
}

make_ids <- function(i, condition) {
  list(
    imageID   = factor(paste0(condition, "_img_", i)),
    subject   = factor(i),
    condition = factor(condition)
  )
}

finalise_cellData <- function(cellData, ids) {
  cellData %>%
    mutate(
      imageID   = ids$imageID,
      subject   = ids$subject,
      condition = ids$condition,
      cellType_sub = factor(cellType_sub, levels = c("CD4","CD8","Treg","B","NK")),
      cellType_major = factor(cellType_major, levels = c("T","B","NK"))
    )
}

safe_bw_diggle <- function(pp, hmax = 2000) {
  bw.diggle(pp)
}

relabel_condition <- function(out, condition_name, i) {
  out$condition <- factor(condition_name)
  out$imageID   <- factor(paste0(condition_name, "_img_", i))
  out
}

# -----------------------------
# Shared helpers you call later
# -----------------------------
allocate_T_subtypes <- function(n_T) {
  t_sub_props <- c(CD4 = 0.70, CD8 = 0.20, Treg = 0.10)
  n_sub <- as.vector(rmultinom(1, n_T, prob = t_sub_props))
  names(n_sub) <- names(t_sub_props)
  n_sub
}

simulate_T_baseline <- function(n_T, win = WIN) {
  t_pts <- simulate_pp(n_T, win)
  pp_T  <- ppp(t_pts$x, t_pts$y, window = win)
  kT    <- density.ppp(pp_T, sigma = safe_bw_diggle(pp_T))
  list(t_pts = t_pts, kT = kT)
}

# -----------------------------
# Extract leaf-level vectors from spicy objects (aligned + canonicalized)
# -----------------------------
extract_leaf_stats <- function(spicy_sub, conditionB, eps = 1e-12) {
  colB <- paste0("condition", conditionB)

  stopifnot(colB %in% colnames(spicy_sub$p.value))
  stopifnot(colB %in% colnames(spicy_sub$coefficient))

  p_raw <- spicy_sub$p.value[, colB]
  c_raw <- spicy_sub$coefficient[, colB]
  inter <- rownames(spicy_sub$p.value)

  df <- tibble(
    interaction = inter,
    leaf = canonical_pair(interaction),
    pval = as.numeric(p_raw),
    coef = as.numeric(c_raw)
  ) %>%
    group_by(leaf) %>%
    slice(1) %>%          # drops NK__CD4 vs CD4__NK duplicates
    ungroup() %>%
    mutate(pval = pmin(pmax(pval, eps), 1 - eps))

  df
}

extract_major_pvals <- function(spicy_major, conditionB) {
  colB <- paste0("condition", conditionB)

  stopifnot(colB %in% colnames(spicy_major$p.value))

  inter <- rownames(spicy_major$p.value)

  tibble(
    interaction = inter,
    major = canonical_pair(inter),
    p_major = as.numeric(spicy_major$p.value[, colB])
  ) %>%
    group_by(major) %>%
    slice(1) %>%
    ungroup()
}

```

# Simulation Functions

## T / CD4 Attracted to NK

### Leaf-level (CD4 is attracted to NK cells)

- NK is generated first, and kNK is estimated from NK points

- baseline T intensity kT is estimated

- CD8 and Treg are sampled from kT (baseline)

- CD4 is sampled from mixture between kT and kNK

```{r}
simulate_leaf_T_to_NK <- function(i, win = WIN, total_cells = 500, w_nk = 0.70) {
  # set.seed(i)

  props <- c(T=0.75, B=0.15, NK=0.10)
  n_T  <- rpois(1, total_cells * props["T"])
  n_B  <- rpois(1, total_cells * props["B"])
  n_NK <- rpois(1, total_cells * props["NK"])
  n_sub <- allocate_T_subtypes(n_T)

  # NK baseline intensity
  nk_pts <- simulate_pp(n_NK, win)
  pp_NK  <- ppp(nk_pts$x, nk_pts$y, window = win)
  kNK    <- density.ppp(pp_NK, sigma = safe_bw_diggle(pp_NK))

  # T baseline intensity
  Tb <- simulate_T_baseline(n_T, win)
  kT <- Tb$kT

  # CD8/Treg from baseline T
  sim_T <- function(n, label) {
    if (n <= 0) return(NULL)
    pp <- rpoint(n, f = kT, win = win)
    tibble(x=pp$x, y=pp$y, cellType_sub=label, cellType_major="T")
  }
  cd8  <- sim_T(n_sub["CD8"],  "CD8")
  treg <- sim_T(n_sub["Treg"], "Treg")

  # CD4 attracted to NK: mix kernel (T + NK), sample CD4 from it
  k_cd4 <- (1 - w_nk) * normalize_im(kT) + w_nk * normalize_im(kNK)
  pp_cd4 <- rpoint(n_sub["CD4"], f = k_cd4, win = win)
  cd4 <- tibble(x=pp_cd4$x, y=pp_cd4$y, cellType_sub="CD4", cellType_major="T")

  b <- simulate_pp(n_B, win) %>% transmute(x,y, cellType_sub="B", cellType_major="B")
  nk <- nk_pts %>% transmute(x,y, cellType_sub="NK", cellType_major="NK")

  cellData <- bind_rows(cd4, cd8, treg, b, nk)
  finalise_cellData(cellData, make_ids(i, "LeafSignal_TtoNK"))
}
```

## Internal-level (T-cells moves towards NK)

- NK is generated first, and kNK is estimated from NK points

- baseline T intensity kT is estimated

- CD4, CD8, and Treg are sampled from a mixture of kT (baseline) and kNK


```{r}
simulate_internal_T_to_NK <- function(i, win = WIN, total_cells = 500, w_nk = 0.70) {
  # set.seed(i)

  props <- c(T=0.75, B=0.15, NK=0.10)
  n_T  <- rpois(1, total_cells * props["T"])
  n_B  <- rpois(1, total_cells * props["B"])
  n_NK <- rpois(1, total_cells * props["NK"])
  n_sub <- allocate_T_subtypes(n_T)

  # NK baseline intensity
  nk_pts <- simulate_pp(n_NK, win)
  pp_NK  <- ppp(nk_pts$x, nk_pts$y, window = win)
  kNK    <- density.ppp(pp_NK, sigma = safe_bw_diggle(pp_NK))

  # T baseline intensity
  Tb <- simulate_T_baseline(n_T, win)
  kT <- Tb$kT

  # All T subtypes attracted to NK
  k_mix <- (1 - w_nk) * normalize_im(kT) + w_nk * normalize_im(kNK)
  sim_mix <- function(n, label) {
    if (n <= 0) return(NULL)
    pp <- rpoint(n, f = k_mix, win = win)
    tibble(x=pp$x, y=pp$y, cellType_sub=label, cellType_major="T")
  }
  t_sub <- bind_rows(
    sim_mix(n_sub["CD4"],  "CD4"),
    sim_mix(n_sub["CD8"],  "CD8"),
    sim_mix(n_sub["Treg"], "Treg")
  )

  b <- simulate_pp(n_B, win) %>% transmute(x,y, cellType_sub="B", cellType_major="B")
  nk <- nk_pts %>% transmute(x,y, cellType_sub="NK", cellType_major="NK")

  cellData <- bind_rows(t_sub, b, nk)
  finalise_cellData(cellData, make_ids(i, "InternalSignal_TtoNK"))
}
```

## NK attracted to CD4 / T cells

### Leaf-Level (NK moves towards CD4)

- T subtypes are generated from baseline T intensity kT (including CD4)

- from the simulated CD4 points, you estimate kCD4

- NK baseline kNK is estimated from random NK points

- NK is sampled from mixture of kNK and kCD4

```{r}
simulate_leaf_NK_to_CD4 <- function(i, win = WIN, total_cells = 500, w_cd4 = 0.70) {
  # set.seed(i)

  props <- c(T=0.75, B=0.15, NK=0.10)
  n_T  <- rpois(1, total_cells * props["T"])
  n_B  <- rpois(1, total_cells * props["B"])
  n_NK <- rpois(1, total_cells * props["NK"])
  n_sub <- allocate_T_subtypes(n_T)

  # T baseline intensity
  Tb <- simulate_T_baseline(n_T, win)
  kT <- Tb$kT

  # Simulate all T subtypes from baseline T (including CD4)
  sim_T <- function(n, label) {
    if (n <= 0) return(NULL)
    pp <- rpoint(n, f = kT, win = win)
    tibble(x=pp$x, y=pp$y, cellType_sub=label, cellType_major="T")
  }
  cd4  <- sim_T(n_sub["CD4"],  "CD4")
  cd8  <- sim_T(n_sub["CD8"],  "CD8")
  treg <- sim_T(n_sub["Treg"], "Treg")

  # CD4 density (attractor target for NK)
  pp_CD4 <- ppp(cd4$x, cd4$y, window = win)
  kCD4 <- density.ppp(pp_CD4, sigma = safe_bw_diggle(pp_CD4))

  # NK baseline density (for mixing)
  nk_base <- simulate_pp(n_NK, win)
  pp_NK   <- ppp(nk_base$x, nk_base$y, window = win)
  kNK     <- density.ppp(pp_NK, sigma = safe_bw_diggle(pp_NK))

  # NK attracted to CD4
  k_nk <- (1 - w_cd4) * normalize_im(kNK) + w_cd4 * normalize_im(kCD4)
  pp_nk <- rpoint(n_NK, f = k_nk, win = win)
  nk <- tibble(x=pp_nk$x, y=pp_nk$y, cellType_sub="NK", cellType_major="NK")

  b <- simulate_pp(n_B, win) %>% transmute(x,y, cellType_sub="B", cellType_major="B")

  cellData <- bind_rows(cd4, cd8, treg, b, nk)
  finalise_cellData(cellData, make_ids(i, "LeafSignal_NKtoCD4"))
}
```

### Internal (NK Moves towards T Cells)

- T subtypes are generated from baseline T intensity kT (including CD4)

- from the simulated CD4 points, you estimate kCD4

- NK baseline kNK is estimated from random NK points

- NK is sampled from mixture of kNK and kT

```{r}
simulate_internal_NK_to_T <- function(i, win = WIN, total_cells = 500, w_t = 0.70) {
  # set.seed(i)

  props <- c(T=0.75, B=0.15, NK=0.10)
  n_T  <- rpois(1, total_cells * props["T"])
  n_B  <- rpois(1, total_cells * props["B"])
  n_NK <- rpois(1, total_cells * props["NK"])
  n_sub <- allocate_T_subtypes(n_T)

  # T baseline intensity
  Tb <- simulate_T_baseline(n_T, win)
  kT <- Tb$kT

  # Simulate all T subtypes from baseline T
  sim_T <- function(n, label) {
    if (n <= 0) return(NULL)
    pp <- rpoint(n, f = kT, win = win)
    tibble(x=pp$x, y=pp$y, cellType_sub=label, cellType_major="T")
  }
  t_sub <- bind_rows(
    sim_T(n_sub["CD4"],  "CD4"),
    sim_T(n_sub["CD8"],  "CD8"),
    sim_T(n_sub["Treg"], "Treg")
  )

  # NK baseline density (for mixing)
  nk_base <- simulate_pp(n_NK, win)
  pp_NK   <- ppp(nk_base$x, nk_base$y, window = win)
  kNK     <- density.ppp(pp_NK, sigma = safe_bw_diggle(pp_NK))

  # NK attracted to overall T intensity
  k_nk <- (1 - w_t) * normalize_im(kNK) + w_t * normalize_im(kT)
  pp_nk <- rpoint(n_NK, f = k_nk, win = win)
  nk <- tibble(x=pp_nk$x, y=pp_nk$y, cellType_sub="NK", cellType_major="NK")

  b <- simulate_pp(n_B, win) %>% transmute(x,y, cellType_sub="B", cellType_major="B")

  cellData <- bind_rows(t_sub, b, nk)
  finalise_cellData(cellData, make_ids(i, "InternalSignal_NKtoT"))
}
```

## Simulate Null

- T subtypes all from baseline kT

- B and NK are uniform Poisson

- No mixing was seen

```{r}
simulate_null <- function(i, win = WIN, total_cells = 500) {
  # set.seed(i)

  props <- c(T=0.75, B=0.15, NK=0.10)
  n_T  <- rpois(1, total_cells * props["T"])
  n_B  <- rpois(1, total_cells * props["B"])
  n_NK <- rpois(1, total_cells * props["NK"])
  n_sub <- allocate_T_subtypes(n_T)

  # baseline T intensity
  Tb <- simulate_T_baseline(n_T, win)
  kT <- Tb$kT

  sim_T <- function(n, label) {
    if (n <= 0) return(NULL)
    pp <- rpoint(n, f = kT, win = win)
    tibble(x=pp$x, y=pp$y, cellType_sub=label, cellType_major="T")
  }

  t_sub <- bind_rows(
    sim_T(n_sub["CD4"],  "CD4"),
    sim_T(n_sub["CD8"],  "CD8"),
    sim_T(n_sub["Treg"], "Treg")
  )

  b  <- simulate_pp(n_B,  win) %>% transmute(x,y, cellType_sub="B",  cellType_major="B")
  nk <- simulate_pp(n_NK, win) %>% transmute(x,y, cellType_sub="NK", cellType_major="NK")

  cellData <- bind_rows(t_sub, b, nk)
  finalise_cellData(cellData, make_ids(i, "Null"))
}
```


# Generate the Dataset

## Creating Wrappers for different spatial strength

This is where we have null vs diferent spatial signal strength

### T / CD4 -> NK

#### CD4 -> NK

```{r}
simulate_leaf_T_to_NK_weak <- function(i, win = WIN, total_cells = 500) {
  out <- simulate_leaf_T_to_NK(i, win = win, total_cells = total_cells, w_nk = 0.30)
  out$condition <- factor("LeafSignal_TtoNK_weak")
  out$imageID   <- factor(paste0("LeafSignal_TtoNK_weak_img_", i))
  out
}

simulate_leaf_T_to_NK_medium <- function(i, win = WIN, total_cells = 500) {
  out <- simulate_leaf_T_to_NK(i, win = win, total_cells = total_cells, w_nk = 0.50)
  out$condition <- factor("LeafSignal_TtoNK_medium")
  out$imageID   <- factor(paste0("LeafSignal_TtoNK_medium_img_", i))
  out
}

simulate_leaf_T_to_NK_strong <- function(i, win = WIN, total_cells = 500) {
  out <- simulate_leaf_T_to_NK(i, win = win, total_cells = total_cells, w_nk = 0.80)
  out$condition <- factor("LeafSignal_TtoNK_strong")
  out$imageID   <- factor(paste0("LeafSignal_TtoNK_strong_img_", i))
  out
}
```

#### T -> NK

```{r}
simulate_internal_T_to_NK_weak <- function(i, win = WIN, total_cells = 500) {
  out <- simulate_internal_T_to_NK(i, win = win, total_cells = total_cells, w_nk = 0.30)
  out$condition <- factor("InternalSignal_TtoNK_weak")
  out$imageID   <- factor(paste0("InternalSignal_TtoNK_weak_img_", i))
  out
}

simulate_internal_T_to_NK_medium <- function(i, win = WIN, total_cells = 500) {
  out <- simulate_internal_T_to_NK(i, win = win, total_cells = total_cells, w_nk = 0.50)
  out$condition <- factor("InternalSignal_TtoNK_medium")
  out$imageID   <- factor(paste0("InternalSignal_TtoNK_medium_img_", i))
  out
}

simulate_internal_T_to_NK_strong <- function(i, win = WIN, total_cells = 500) {
  out <- simulate_internal_T_to_NK(i, win = win, total_cells = total_cells, w_nk = 0.80)
  out$condition <- factor("InternalSignal_TtoNK_strong")
  out$imageID   <- factor(paste0("InternalSignal_TtoNK_strong_img_", i))
  out
}
```

### NK -> CD4 / T

#### NK -> CD4

```{r}
simulate_leaf_NK_to_CD4_weak <- function(i, win = WIN, total_cells = 500) {
  out <- simulate_leaf_NK_to_CD4(i, win = win, total_cells = total_cells, w_cd4 = 0.30)
  out$condition <- factor("LeafSignal_NKtoCD4_weak")
  out$imageID   <- factor(paste0("LeafSignal_NKtoCD4_weak_img_", i))
  out
}

simulate_leaf_NK_to_CD4_medium <- function(i, win = WIN, total_cells = 500) {
  out <- simulate_leaf_NK_to_CD4(i, win = win, total_cells = total_cells, w_cd4 = 0.50)
  out$condition <- factor("LeafSignal_NKtoCD4_medium")
  out$imageID   <- factor(paste0("LeafSignal_NKtoCD4_medium_img_", i))
  out
}

simulate_leaf_NK_to_CD4_strong <- function(i, win = WIN, total_cells = 500) {
  out <- simulate_leaf_NK_to_CD4(i, win = win, total_cells = total_cells, w_cd4 = 0.80)
  out$condition <- factor("LeafSignal_NKtoCD4_strong")
  out$imageID   <- factor(paste0("LeafSignal_NKtoCD4_strong_img_", i))
  out
}
```

#### NK -> T

```{r}
simulate_internal_NK_to_T_weak <- function(i, win = WIN, total_cells = 500) {
  out <- simulate_internal_NK_to_T(i, win = win, total_cells = total_cells, w_t = 0.30)
  out$condition <- factor("InternalSignal_NKtoT_weak")
  out$imageID   <- factor(paste0("InternalSignal_NKtoT_weak_img_", i))
  out
}

simulate_internal_NK_to_T_medium <- function(i, win = WIN, total_cells = 500) {
  out <- simulate_internal_NK_to_T(i, win = win, total_cells = total_cells, w_t = 0.50)
  out$condition <- factor("InternalSignal_NKtoT_medium")
  out$imageID   <- factor(paste0("InternalSignal_NKtoT_medium_img_", i))
  out
}

simulate_internal_NK_to_T_strong <- function(i, win = WIN, total_cells = 500) {
  out <- simulate_internal_NK_to_T(i, win = win, total_cells = total_cells, w_t = 0.80)
  out$condition <- factor("InternalSignal_NKtoT_strong")
  out$imageID   <- factor(paste0("InternalSignal_NKtoT_strong_img_", i))
  out
}
```

## Generating Mixed Control and Treatment 

### Leaf T / CD4 -> NK

#### Control

```{r}
simulate_leaf_T_to_NK_control <- function(
  i,
  control_strength = 0.20,
  treatment_strength = 0.50,
  win = WIN,
  total_cells = 500
) {
  out <- simulate_leaf_T_to_NK(
    i,
    win = win,
    total_cells = total_cells,
    w_nk = control_strength
  )

  condition_name <- paste0(
    "Control_Leaf_TtoNK_",
    gsub("\\.", "p", as.character(control_strength))
  )

  relabel_condition(out, condition_name, i)
}
```

#### Treatment

```{r}
simulate_leaf_T_to_NK_treatment <- function(
  i,
  control_strength = 0.20,
  treatment_strength = 0.50,
  win = WIN,
  total_cells = 500
) {
  out <- simulate_leaf_T_to_NK(
    i,
    win = win,
    total_cells = total_cells,
    w_nk = treatment_strength
  )

  condition_name <- paste0(
    "Treatment_Leaf_TtoNK_",
    gsub("\\.", "p", as.character(treatment_strength))
  )

  relabel_condition(out, condition_name, i)
}
```

### Internal T -> NK

#### Control

```{r}
simulate_internal_T_to_NK_control <- function(
  i,
  control_strength = 0.20,
  treatment_strength = 0.50,
  win = WIN,
  total_cells = 500
) {
  out <- simulate_internal_T_to_NK(
    i,
    win = win,
    total_cells = total_cells,
    w_nk = control_strength
  )

  condition_name <- paste0(
    "Control_Internal_TtoNK_",
    gsub("\\.", "p", as.character(control_strength))
  )

  relabel_condition(out, condition_name, i)
}
```

#### Treatment

```{r}
simulate_internal_T_to_NK_treatment <- function(
  i,
  control_strength = 0.20,
  treatment_strength = 0.50,
  win = WIN,
  total_cells = 500
) {
  out <- simulate_internal_T_to_NK(
    i,
    win = win,
    total_cells = total_cells,
    w_nk = treatment_strength
  )

  condition_name <- paste0(
    "Treatment_Internal_TtoNK_",
    gsub("\\.", "p", as.character(treatment_strength))
  )

  relabel_condition(out, condition_name, i)
}
```

### Leaf NK -> CD4

#### Control

```{r}
simulate_leaf_NK_to_CD4_control <- function(
  i,
  control_strength = 0.20,
  treatment_strength = 0.50,
  win = WIN,
  total_cells = 500
) {
  out <- simulate_leaf_NK_to_CD4(
    i,
    win = win,
    total_cells = total_cells,
    w_cd4 = control_strength
  )

  condition_name <- paste0(
    "Control_Leaf_NKtoCD4_",
    gsub("\\.", "p", as.character(control_strength))
  )

  relabel_condition(out, condition_name, i)
}
```

#### Treatment

```{r}
simulate_leaf_NK_to_CD4_treatment <- function(
  i,
  control_strength = 0.20,
  treatment_strength = 0.50,
  win = WIN,
  total_cells = 500
) {
  out <- simulate_leaf_NK_to_CD4(
    i,
    win = win,
    total_cells = total_cells,
    w_cd4 = treatment_strength
  )

  condition_name <- paste0(
    "Treatment_Leaf_NKtoCD4_",
    gsub("\\.", "p", as.character(treatment_strength))
  )

  relabel_condition(out, condition_name, i)
}
```

### Internal NK -> T

#### Control

```{r}
simulate_internal_NK_to_T_control <- function(
  i,
  control_strength = 0.20,
  treatment_strength = 0.50,
  win = WIN,
  total_cells = 500
) {
  out <- simulate_internal_NK_to_T(
    i,
    win = win,
    total_cells = total_cells,
    w_t = control_strength
  )

  condition_name <- paste0(
    "Control_Internal_NKtoT_",
    gsub("\\.", "p", as.character(control_strength))
  )

  relabel_condition(out, condition_name, i)
}
```

#### Treatment

```{r}
simulate_internal_NK_to_T_treatment <- function(
  i,
  control_strength = 0.20,
  treatment_strength = 0.50,
  win = WIN,
  total_cells = 500
) {
  out <- simulate_internal_NK_to_T(
    i,
    win = win,
    total_cells = total_cells,
    w_t = treatment_strength
  )

  condition_name <- paste0(
    "Treatment_Internal_NKtoT_",
    gsub("\\.", "p", as.character(treatment_strength))
  )

  relabel_condition(out, condition_name, i)
}
```

## Running the Simulation

### Running the Null Vs. Various Spatial Strength

```{r}
N_PER_COND <- 100

# cellData_all <- bind_rows(
#   lapply(1:N_PER_COND, simulate_null),
# 
#   # Part A: T/CD4 -> NK
#   lapply(1:N_PER_COND, simulate_leaf_T_to_NK),
#   lapply(1:N_PER_COND, simulate_internal_T_to_NK),
# 
#   # Part B: NK -> CD4/T
#   lapply(1:N_PER_COND, simulate_leaf_NK_to_CD4),
#   lapply(1:N_PER_COND, simulate_internal_NK_to_T)
# )

cellData_all <- bind_rows(
  lapply(1:N_PER_COND, simulate_null),

  # -----------------------------
  # Part A: T/CD4 -> NK
  # -----------------------------
  lapply(1:N_PER_COND, simulate_leaf_T_to_NK_weak),
  lapply(1:N_PER_COND, simulate_leaf_T_to_NK_medium),
  lapply(1:N_PER_COND, simulate_leaf_T_to_NK_strong),

  lapply(1:N_PER_COND, simulate_internal_T_to_NK_weak),
  lapply(1:N_PER_COND, simulate_internal_T_to_NK_medium),
  lapply(1:N_PER_COND, simulate_internal_T_to_NK_strong),

  # -----------------------------
  # Part B: NK -> CD4 / T
  # -----------------------------
  lapply(1:N_PER_COND, simulate_leaf_NK_to_CD4_weak),
  lapply(1:N_PER_COND, simulate_leaf_NK_to_CD4_medium),
  lapply(1:N_PER_COND, simulate_leaf_NK_to_CD4_strong),

  lapply(1:N_PER_COND, simulate_internal_NK_to_T_weak),
  lapply(1:N_PER_COND, simulate_internal_NK_to_T_medium),
  lapply(1:N_PER_COND, simulate_internal_NK_to_T_strong)
)

table(cellData_all$condition)
```

#### Verifying it works

```{r}
# one row per image-condition
img_summary <- cellData_all %>%
  group_by(condition, imageID) %>%
  summarise(
    n_cells = n(),
    n_CD4   = sum(cellType_sub == "CD4"),
    n_CD8   = sum(cellType_sub == "CD8"),
    n_Treg  = sum(cellType_sub == "Treg"),
    n_B     = sum(cellType_sub == "B"),
    n_NK    = sum(cellType_sub == "NK"),
    xmin = min(x), xmax = max(x),
    ymin = min(y), ymax = max(y),
    .groups = "drop"
  )

img_summary %>%
  group_by(condition) %>%
  summarise(
    n_images = n(),
    mean_cells = mean(n_cells),
    mean_CD4 = mean(n_CD4),
    mean_NK  = mean(n_NK)
  )

plot_image <- function(dat, img) {
  dat %>%
    filter(imageID == img) %>%
    ggplot(aes(x, y, color = cellType_sub)) +
    geom_point(size = 1, alpha = 0.8) +
    coord_fixed() +
    theme_bw() +
    ggtitle(as.character(img))
}

# example
some_imgs <- cellData_all %>%
  distinct(condition, imageID) %>%
  group_by(condition) %>%
  dplyr::slice(1) %>%
  ungroup()

plot_image(cellData_all, some_imgs$imageID[1])
```

```{r}
mean_nn_dist <- function(dat, from_type, to_type, major = FALSE) {
  a <- if (major) dat %>% filter(cellType_major == from_type) else dat %>% filter(cellType_sub == from_type)
  b <- if (major) dat %>% filter(cellType_major == to_type) else dat %>% filter(cellType_sub == to_type)

  if (nrow(a) == 0 || nrow(b) == 0) return(NA_real_)

  XA <- cbind(a$x, a$y)
  XB <- cbind(b$x, b$y)

  dmat <- as.matrix(dist(rbind(XA, XB)))
  na <- nrow(XA)
  nb <- nrow(XB)

  dAB <- dmat[1:na, (na + 1):(na + nb), drop = FALSE]
  mean(apply(dAB, 1, min))
}

nn_summary <- cellData_all %>%
  group_by(condition, imageID) %>%
  group_modify(~{
    dat <- .x
    tibble(
      CD4_to_NK = mean_nn_dist(dat, "CD4", "NK"),
      NK_to_CD4 = mean_nn_dist(dat, "NK", "CD4"),
      NK_to_T   = mean_nn_dist(dat, "NK", "T", major = TRUE),
      CD8_to_NK = mean_nn_dist(dat, "CD8", "NK"),
      Treg_to_NK = mean_nn_dist(dat, "Treg", "NK")
    )
  }) %>%
  ungroup()

nn_summary %>%
  group_by(condition) %>%
  summarise(across(c(CD4_to_NK, NK_to_CD4, NK_to_T, CD8_to_NK, Treg_to_NK), mean, na.rm = TRUE))

nn_summary %>%
  filter(str_detect(condition, "LeafSignal_TtoNK")) %>%
  mutate(strength = case_when(
    str_detect(condition, "weak") ~ "weak",
    str_detect(condition, "medium") ~ "medium",
    str_detect(condition, "strong") ~ "strong"
  )) %>%
  group_by(strength) %>%
  summarise(mean_CD4_to_NK = mean(CD4_to_NK, na.rm = TRUE))
```


### Mixed Spatial Strength in Treatment and Control

```{r}
N_PER_COND <- 100

cellData_strength_pairs <- bind_rows(
  lapply(1:N_PER_COND, simulate_leaf_T_to_NK_control,
         control_strength = 0.20, treatment_strength = 0.50),
  lapply(1:N_PER_COND, simulate_leaf_T_to_NK_treatment,
         control_strength = 0.20, treatment_strength = 0.50),

  lapply(1:N_PER_COND, simulate_internal_T_to_NK_control,
         control_strength = 0.20, treatment_strength = 0.50),
  lapply(1:N_PER_COND, simulate_internal_T_to_NK_treatment,
         control_strength = 0.20, treatment_strength = 0.50),

  lapply(1:N_PER_COND, simulate_leaf_NK_to_CD4_control,
         control_strength = 0.20, treatment_strength = 0.50),
  lapply(1:N_PER_COND, simulate_leaf_NK_to_CD4_treatment,
         control_strength = 0.20, treatment_strength = 0.50),

  lapply(1:N_PER_COND, simulate_internal_NK_to_T_control,
         control_strength = 0.20, treatment_strength = 0.50),
  lapply(1:N_PER_COND, simulate_internal_NK_to_T_treatment,
         control_strength = 0.20, treatment_strength = 0.50)
)

table(cellData_strength_pairs$condition)
```

#### Verifying it all works

```{r}
# one row per image-condition
img_summary_pairs <- cellData_strength_pairs %>%
  group_by(condition, imageID) %>%
  summarise(
    n_cells = n(),
    n_CD4   = sum(cellType_sub == "CD4"),
    n_CD8   = sum(cellType_sub == "CD8"),
    n_Treg  = sum(cellType_sub == "Treg"),
    n_B     = sum(cellType_sub == "B"),
    n_NK    = sum(cellType_sub == "NK"),
    xmin = min(x), xmax = max(x),
    ymin = min(y), ymax = max(y),
    .groups = "drop"
  )

# quick sanity check by condition
img_summary_pairs %>%
  group_by(condition) %>%
  summarise(
    n_images = n(),
    mean_cells = mean(n_cells),
    mean_CD4 = mean(n_CD4),
    mean_NK  = mean(n_NK),
    .groups = "drop"
  )

# plotting helper
plot_image <- function(dat, img) {
  dat %>%
    filter(imageID == img) %>%
    ggplot(aes(x, y, color = cellType_sub)) +
    geom_point(size = 1, alpha = 0.8) +
    coord_fixed() +
    theme_bw() +
    ggtitle(as.character(img))
}

# plot one example image from each condition
some_imgs_pairs <- cellData_strength_pairs %>%
  distinct(condition, imageID) %>%
  group_by(condition) %>%
  dplyr::slice(1) %>%
  ungroup()

plot_image(cellData_strength_pairs, some_imgs_pairs$imageID[1])

# nearest-neighbour distance helper
mean_nn_dist <- function(dat, from_type, to_type, major = FALSE) {
  a <- if (major) dat %>% filter(cellType_major == from_type) else dat %>% filter(cellType_sub == from_type)
  b <- if (major) dat %>% filter(cellType_major == to_type) else dat %>% filter(cellType_sub == to_type)

  if (nrow(a) == 0 || nrow(b) == 0) return(NA_real_)

  XA <- cbind(a$x, a$y)
  XB <- cbind(b$x, b$y)

  dmat <- as.matrix(dist(rbind(XA, XB)))
  na <- nrow(XA)
  nb <- nrow(XB)

  dAB <- dmat[1:na, (na + 1):(na + nb), drop = FALSE]
  mean(apply(dAB, 1, min))
}

# per-image NN summaries
nn_summary_pairs <- cellData_strength_pairs %>%
  group_by(condition, imageID) %>%
  group_modify(~{
    dat <- .x
    tibble(
      CD4_to_NK  = mean_nn_dist(dat, "CD4", "NK"),
      NK_to_CD4  = mean_nn_dist(dat, "NK", "CD4"),
      NK_to_T    = mean_nn_dist(dat, "NK", "T", major = TRUE),
      CD8_to_NK  = mean_nn_dist(dat, "CD8", "NK"),
      Treg_to_NK = mean_nn_dist(dat, "Treg", "NK")
    )
  }) %>%
  ungroup()

# average NN distances by condition
nn_summary_pairs %>%
  group_by(condition) %>%
  summarise(
    across(c(CD4_to_NK, NK_to_CD4, NK_to_T, CD8_to_NK, Treg_to_NK), mean, na.rm = TRUE),
    .groups = "drop"
  )
```


```{r}
nn_summary_pairs_labeled <- nn_summary_pairs %>%
  mutate(
    scenario = case_when(
      str_detect(condition, "Leaf_TtoNK")     ~ "Leaf_TtoNK",
      str_detect(condition, "Internal_TtoNK") ~ "Internal_TtoNK",
      str_detect(condition, "Leaf_NKtoCD4")   ~ "Leaf_NKtoCD4",
      str_detect(condition, "Internal_NKtoT") ~ "Internal_NKtoT"
    ),
    group = case_when(
      str_detect(condition, "^Control")   ~ "Control",
      str_detect(condition, "^Treatment") ~ "Treatment"
    ),
    strength = str_extract(condition, "[0-9]p[0-9]+") %>%
      str_replace("p", ".") %>%
      as.numeric()
  )

nn_summary_pairs_labeled %>%
  group_by(scenario, group, strength) %>%
  summarise(
    mean_CD4_to_NK = mean(CD4_to_NK, na.rm = TRUE),
    mean_NK_to_CD4 = mean(NK_to_CD4, na.rm = TRUE),
    mean_NK_to_T   = mean(NK_to_T, na.rm = TRUE),
    .groups = "drop"
  )
```

# Building the SpatialExperiment

```{r}
spe <- SpatialExperiment(
  spatialCoords = as.matrix(cellData_all[, c("x", "y")]),
  colData = cellData_all %>%
    select(imageID, subject, condition, cellType_sub, cellType_major)
)

spe_pairs <- SpatialExperiment(
  spatialCoords = as.matrix(cellData_strength_pairs[, c("x", "y")]),
  colData = cellData_strength_pairs %>%
    select(imageID, subject, condition, cellType_sub, cellType_major)
)
```

```{r}
saveRDS(spe, "spe.rds")
saveRDS(spe_pairs, "spe_pairs.rds")
```


# running spicyR: Null vs each condition

```{r}
run_spicy_contrast <- function(spe, cond_A = "Null", cond_B, level = c("sub", "major")) {
  level <- match.arg(level)
  cell_col <- if (level == "sub") "cellType_sub" else "cellType_major"

  spe2 <- spe[, spe$condition %in% c(cond_A, cond_B)]
  spe2$condition <- droplevels(spe2$condition)
  spe2$imageID   <- droplevels(spe2$imageID)
  spe2[[cell_col]] <- droplevels(spe2[[cell_col]])

  spicy(
    spe2,
    condition = "condition",
    subject   = "subject",
    cellType  = cell_col,
    imageID   = "imageID"
  )
}
```

# Running the Simulation

```{r}
conds <- c(
  "LeafSignal_TtoNK_weak",    "LeafSignal_TtoNK_medium",    "LeafSignal_TtoNK_strong",
  "InternalSignal_TtoNK_weak","InternalSignal_TtoNK_medium","InternalSignal_TtoNK_strong",
  "LeafSignal_NKtoCD4_weak",  "LeafSignal_NKtoCD4_medium",  "LeafSignal_NKtoCD4_strong",
  "InternalSignal_NKtoT_weak","InternalSignal_NKtoT_medium","InternalSignal_NKtoT_strong"
)

spicy_sub_list   <- setNames(lapply(conds, \(cc) run_spicy_contrast(spe, "Null", cc, "sub")),   conds)
spicy_major_list <- setNames(lapply(conds, \(cc) run_spicy_contrast(spe, "Null", cc, "major")), conds)

# Quick sanity: top pairs
lapply(spicy_sub_list, topPairs)
lapply(spicy_major_list, topPairs)
```

```{r}
example_imgs <- cellData_all %>%
  distinct(imageID, condition) %>%
  group_by(condition) %>%
  slice_sample(n = 1) %>%
  pull(imageID)

cellData_all %>%
  filter(imageID %in% example_imgs) %>%
  ggplot(aes(x, y, color = cellType_sub)) +
  geom_point(size = 0.6, alpha = 0.7) +
  coord_fixed() +
  facet_wrap(~ condition) +
  theme_minimal()
```

# TreeBH

TreeBH needs ONLY:
  
  1. Leaf p-values -> numeric vector
  2. Tree structure -> integer matrix
  
TreeBH assumes that you already know the correct biological hierarchy and asks: "Given this tree, how do I control FDR?"

- Internal leaves are implicitly recomputed via Simes from the leaves

What is Simes? It's a way to combine several p-values into a single-value or the hypothesis "at least one of these is non-null"

TreeBH instead of controlling for global FDR, it tries to control something else called sFDR where instead of controllling FDR for all the hypotheses, its conly controlling the FDR of all the hypothesis chosen and their children

So how are we going to transform the tests we have into a tree?

- Level 0: root 
- Level 1: Major-Major interaction family
  - Group by the major cell types of the pair:
    - `T__T`
    - `T__B`
    - `T__NK`
    - `B__B`
    - `B__NK`
    - `NK__NK`
- Level 2: Subtype-Subtype within the family
  - Under `T__T`
    - `CD4__CD4`
    - `CD4__CD8`
    - `CD4__Treg`
    - `CD8__CD8`
    - `CD8__Treg`
    - `Treg__Treg`
  - Under `T__B`
    - `CD4__B`
    - `CD8__B`
    - `Treg__B`
  - Under `B__B`
    - `B__B`
  - Under `B__NK`
    - `B__NK`
  - Under `NK__NK`
    - `NK__NK`

Note that we have hypothesis such as `CD8__CD4` and `CD4__CD8`, we'll juts keep one of the two

```{r}
# -----------------------------
# TreeBH: build groups + run
# -----------------------------

build_treebh_inputs <- function(spicy_sub, conditionB,
                                major_map = MAJOR_MAP,
                                eps = 1e-12) {
  # 1) leaf-level pvals/coefs (canonicalized, de-duplicated)
  df_leaf <- extract_leaf_stats(spicy_sub, conditionB = conditionB, eps = eps) %>%
    tidyr::separate(leaf, into = c("a", "b"), sep = "__", remove = FALSE) %>%
    dplyr::mutate(
      ma = unname(major_map[a]),
      mb = unname(major_map[b]),
      fam = canonical_pair(paste(ma, mb, sep = "__"))
    ) %>%
    dplyr::select(leaf, fam, pval, coef)

  # 2) TreeBH groups matrix (2 levels)
  #    level1: family index
  #    level2: leaf-within-family index
  groups <- cbind(
    level1 = as.integer(factor(df_leaf$fam)),
    level2 = as.integer(factor(paste(df_leaf$fam, df_leaf$leaf, sep = "/")))
  )

  list(
    df_leaf = df_leaf,
    pvals   = df_leaf$pval,
    groups  = groups
  )
}

run_treebh <- function(spicy_sub, conditionB,
                       q = c(0.2, 0.2),
                       major_map = MAJOR_MAP,
                       eps = 1e-12) {

  inp <- build_treebh_inputs(
    spicy_sub  = spicy_sub,
    conditionB = conditionB,
    major_map  = major_map,
    eps        = eps
  )

  res <- TreeBH::get_TreeBH_selections(
    pvals  = inp$pvals,
    groups = inp$groups,
    q      = q
  )

  # TreeBH returns a matrix with a column per level; leaf selections are in level 2
  leaf_reject <- (res[, 2] == 1)

  inp$df_leaf %>%
    dplyr::mutate(
      reject_treebh = leaf_reject
    ) %>%
    dplyr::arrange(fam, leaf)
}
```

```{r}
q_tree <- c(0.2, 0.2)

treebh_sub_results <- purrr::imap(spicy_sub_list, \(sp_obj, cc) {
  run_treebh(
    spicy_sub  = sp_obj,
    conditionB = cc,      # e.g. "LeafSignal_TtoNK"
    q          = q_tree
  ) %>%
    mutate(condition = cc)
})

treebh_sub_df <- bind_rows(treebh_sub_results)

treebh_sub_df %>%
  group_by(condition) %>%
  summarise(n_reject = sum(reject_treebh), .groups = "drop")
```

# DART

DART composes of two stages:
  1. Building a tree from a distance matrix between hypotheses
  2. Tests bottom-up on the tree
  
Similarly to treeBH, it needs leaf hypotheses 

DART assumes you don't know the hierarchy but have distances that hint at co-status and asks: "Can I learn a tree from distances and then test on it safely?"

- Don't know the correct hierarchy
- Only know pairwise distances between leaf hypotheses
- Distance reflects likelihood for shared signal

In treeBH the internal represents the hypothesis: At least one of my descendant leaf hypothesis is non-null but in DART, the internal node represents: "These leaf hypotheses are close enough (in distance) that they may share signal

```{r}
extract_leaf_for_dart <- function(spicy_sub, conditionB, eps = 1e-10) {
  colB <- paste0("condition", conditionB)

  stopifnot(colB %in% colnames(spicy_sub$p.value))
  stopifnot(colB %in% colnames(spicy_sub$coefficient))

  inter <- rownames(spicy_sub$p.value)

  df <- tibble(
    interaction = inter,
    leaf = canonical_pair(interaction),
    pval = as.numeric(spicy_sub$p.value[, colB]),
    coef = as.numeric(spicy_sub$coefficient[, colB])
  ) %>%
    group_by(leaf) %>%
    slice(1) %>%                     # drop A__B vs B__A duplicates
    ungroup() %>%
    mutate(pval = pmin(pmax(pval, eps), 1 - eps)) %>%
    arrange(leaf)

  list(
    df = df,
    pvals = setNames(df$pval, df$leaf),
    coefs = setNames(df$coef, df$leaf)
  )
}

make_hypothesis_distance <- function(leaf_names, major_map = MAJOR_MAP) {

  cell_dist <- function(x, y) {
    if (x == y) return(0L)
    if (major_map[[x]] == major_map[[y]]) return(1L)
    2L
  }

  parse_pair <- function(s) {
    ab <- strsplit(s, "__", fixed = TRUE)[[1]]
    ab <- sort(ab)
    list(a = ab[1], b = ab[2])
  }

  pairs <- lapply(leaf_names, parse_pair)

  m <- length(leaf_names)
  D <- matrix(0, m, m, dimnames = list(leaf_names, leaf_names))

  for (i in seq_len(m)) {
    a <- pairs[[i]]$a; b <- pairs[[i]]$b
    for (j in seq_len(m)) {
      c <- pairs[[j]]$a; d <- pairs[[j]]$b
      D[i, j] <- min(
        cell_dist(a, c) + cell_dist(b, d),
        cell_dist(a, d) + cell_dist(b, c)
      )
    }
  }

  D
}

run_dart_one <- function(spicy_sub, spe, cond_B,
                         cond_A = "Null",
                         alpha = 0.05,
                         eps = 1e-10,
                         Mgroup = 2) {

  # 1) leaf pvals + coefs
  leaf <- extract_leaf_for_dart(spicy_sub, conditionB = cond_B, eps = eps)
  pvals <- leaf$pvals
  coefs <- leaf$coefs
  leaf_names <- names(pvals)

  # 2) distance matrix (normalize to [0,1])
  D <- make_hypothesis_distance(leaf_names, major_map = MAJOR_MAP)
  D2 <- D / max(D)

  # 3) signed z-scores (two-sided)
  z <- qnorm(1 - as.numeric(pvals) / 2) * sign(as.numeric(coefs))
  T1 <- matrix(z, nrow = 1)  # 1 x m

  # 4) n = number of images for this contrast
  spe2 <- spe[, spe$condition %in% c(cond_A, cond_B)]
  n_images <- length(unique(spe2$imageID))

  m <- length(pvals)
  n <- n_images
  grids <- 16 / sqrt(n * log(m) * log(log(m)))

  # 5) build aggregation tree + test
  Atree <- DART::A.tree.mult(
    grids  = grids,
    Dist0  = D2,
    Mgroup = Mgroup
  )

  res <- DART::test.mult(
    alpha = alpha,
    Llist = Atree$Llist,
    Dist0 = D2,
    T1    = T1
  )

  # leaf indices rejected by DART
  dart_idx <- res[[length(res)]]

  dart_reject <- rep(FALSE, m)
  if (length(dart_idx) > 0) dart_reject[dart_idx] <- TRUE

  # return tidy leaf table + also raw objects for debugging
  list(
    summary = tibble(
      condition = cond_B,
      leaf      = leaf_names,
      pval      = as.numeric(pvals),
      coef      = as.numeric(coefs),
      reject    = dart_reject
    ) %>% arrange(pval),
    dart_idx = dart_idx,
    Atree    = Atree,
    Dist     = D2
  )
}

```

```{r}
dart_out_list <- purrr::imap(spicy_sub_list, \(sp_obj, cc) {
  run_dart_one(
    spicy_sub = sp_obj,
    spe       = spe,
    cond_B    = cc,
    alpha     = 0.05
  )
})

# leaf-level results combined
dart_df <- bind_rows(lapply(dart_out_list, `[[`, "summary"))

# quick counts
dart_df %>%
  group_by(condition) %>%
  summarise(n_reject = sum(reject), .groups = "drop")
```

# P-Filter (Not Working Rn)

```{r}
extract_pfilter_inputs <- function(spicy_sub, spicy_major, cond_B, eps = 1e-12) {
  colB <- paste0("condition", cond_B)

  stopifnot(colB %in% colnames(spicy_sub$p.value))
  stopifnot(colB %in% colnames(spicy_major$p.value))

  # ---- major level (families)
  major_df <- tibble(
    interaction = rownames(spicy_major$p.value),
    fam = canonical_pair(rownames(spicy_major$p.value)),
    p_major = as.numeric(spicy_major$p.value[, colB])
  ) %>%
    group_by(fam) %>%
    slice(1) %>%
    ungroup() %>%
    mutate(p_major = pmin(pmax(p_major, eps), 1 - eps)) %>%
    arrange(fam)

  # ---- leaf level (subtype pairs)
  sub_df <- tibble(
    interaction = rownames(spicy_sub$p.value),
    leaf = canonical_pair(rownames(spicy_sub$p.value)),
    p_sub = as.numeric(spicy_sub$p.value[, colB])
  ) %>%
    group_by(leaf) %>%
    slice(1) %>%                 # drop A__B vs B__A duplicates
    ungroup() %>%
    tidyr::separate(leaf, into = c("a","b"), sep="__", remove = FALSE) %>%
    mutate(
      fam = canonical_pair(paste0(MAJOR_MAP[a], "__", MAJOR_MAP[b])),
      p_sub = pmin(pmax(p_sub, eps), 1 - eps)
    ) %>%
    select(leaf, fam, p_sub) %>%
    arrange(fam, leaf)

  list(major_df = major_df, sub_df = sub_df)
}


run_pfilter_one <- function(spicy_sub, spicy_major, cond_B,
                            alpha_major = 0.05,
                            alpha_leaf  = 0.05,
                            eps = 1e-12) {

  inp <- extract_pfilter_inputs(spicy_sub, spicy_major, cond_B, eps = eps)
  major_df <- inp$major_df
  sub_df   <- inp$sub_df

  n_major <- nrow(major_df)
  n_sub   <- nrow(sub_df)
  n_total <- n_major + n_sub

  # Combine p-values in the order: majors first, then leaves
  P_all <- c(major_df$p_major, sub_df$p_sub)

  # groups matrix: n_total x 2
  groups <- matrix(0L, nrow = n_total, ncol = 2)

  # Layer 1: hierarchy constraint
  # majors: 1..n_major
  groups[1:n_major, 1] <- seq_len(n_major)

  # leaves: inherit parent's major group
  parent_idx <- match(sub_df$fam, major_df$fam)
  if (anyNA(parent_idx)) {
    stop("Some leaf families did not match any major family. Check MAJOR_MAP / canonicalization.")
  }
  groups[(n_major + 1):n_total, 1] <- parent_idx

  # Layer 2: discovery layer (each test unique)
  groups[, 2] <- seq_len(n_total)

  # Run pfilter (from your pfilter.R)
  res <- pfilter(
    P      = P_all,
    alphas = c(alpha_major, alpha_leaf),
    groups = groups
  )

  # Extract leaf rejections (last n_sub)
  leaf_reject <- res[(n_major + 1):n_total] == 1

  # Return tidy table at leaf level
  tibble(
    condition = cond_B,
    leaf      = sub_df$leaf,
    fam       = sub_df$fam,
    pval      = sub_df$p_sub,
    reject    = leaf_reject
  ) %>% arrange(fam, pval)
}
```

```{r}
pfilter_df <- purrr::imap_dfr(spicy_sub_list, \(sp_sub, cc) {
  run_pfilter_one(
    spicy_sub   = sp_sub,
    spicy_major = spicy_major_list[[cc]],
    cond_B      = cc,
    alpha_major = 0.05,
    alpha_leaf  = 0.05
  )
})

# Count discoveries per condition
pfilter_df %>%
  group_by(condition) %>%
  summarise(n_reject = sum(reject), .groups = "drop")
```

# Clustering + Thinning (BY)

```{r}
extract_major_leaf_tables <- function(spicy_sub, spicy_major, cond_B, eps = 1e-12) {
  colB <- paste0("condition", cond_B)

  stopifnot(colB %in% colnames(spicy_sub$p.value))
  stopifnot(colB %in% colnames(spicy_major$p.value))

  # Major families (canonicalized)
  major_df <- tibble(
    interaction = rownames(spicy_major$p.value),
    fam = canonical_pair(rownames(spicy_major$p.value)),
    p_major = as.numeric(spicy_major$p.value[, colB])
  ) %>%
    group_by(fam) %>%
    slice(1) %>%
    ungroup() %>%
    mutate(p_major = pmin(pmax(p_major, eps), 1 - eps)) %>%
    arrange(fam)

  # Leaf hypotheses (canonicalized + mapped to fam via MAJOR_MAP)
  sub_df <- tibble(
    interaction = rownames(spicy_sub$p.value),
    leaf = canonical_pair(rownames(spicy_sub$p.value)),
    p_sub = as.numeric(spicy_sub$p.value[, colB])
  ) %>%
    group_by(leaf) %>%
    slice(1) %>%   # drop CD4__NK vs NK__CD4 duplicates
    ungroup() %>%
    tidyr::separate(leaf, into = c("a", "b"), sep = "__", remove = FALSE) %>%
    mutate(
      fam = canonical_pair(paste0(MAJOR_MAP[a], "__", MAJOR_MAP[b])),
      p_sub = pmin(pmax(p_sub, eps), 1 - eps)
    ) %>%
    select(leaf, fam, p_sub) %>%
    arrange(fam, leaf)

  list(major_df = major_df, sub_df = sub_df)
}

cluster_test_major <- function(p_major, q = 0.05, method = c("BY", "BH")) {
  method <- match.arg(method)

  adj <- p.adjust(p_major, method = method)
  sig <- adj <= q

  sig_fams <- names(p_major)[sig]

  # u1 cutoff = largest *raw* p-value among rejected major clusters (as per your note)
  u1 <- if (any(sig)) max(p_major[sig]) else q

  list(sig_fams = sig_fams, u1 = u1, adj = adj)
}

calculate_trimming_p <- function(z_subtype, u1_cutoff, rho = 0.5, m0_m = 0.9) {
  num <- m0_m * pnorm((qnorm(1 - u1_cutoff) - rho * z_subtype) / sqrt(1 - rho^2))
  den <- (m0_m * u1_cutoff + (1 - m0_m))
  p_trim <- num / den
  pmin(pmax(p_trim, 0), 1)
}

run_cluster_trim_one <- function(spicy_sub, spicy_major, cond_B,
                                 q = 0.05,
                                 major_method = c("BY", "BH"),
                                 rho = 0.5,
                                 m0_m = 0.9,
                                 two_sided_z = FALSE,
                                 eps = 1e-12) {

  major_method <- match.arg(major_method)

  inp <- extract_major_leaf_tables(spicy_sub, spicy_major, cond_B, eps = eps)
  major_df <- inp$major_df
  sub_df   <- inp$sub_df

  p_major <- major_df$p_major
  names(p_major) <- major_df$fam

  # ---- Stage 1: major cluster testing
  maj <- cluster_test_major(p_major, q = q, method = major_method)
  sig_fams <- maj$sig_fams
  u1 <- maj$u1

  # ---- Stage 2: trimming within significant families
  out <- sub_df %>%
    mutate(
      in_sig_family = fam %in% sig_fams,
      z_subtype = if (two_sided_z) qnorm(1 - p_sub / 2) else qnorm(1 - p_sub),
      p_trimmed = ifelse(
        in_sig_family,
        vapply(z_subtype, calculate_trimming_p, numeric(1),
               u1_cutoff = u1, rho = rho, m0_m = m0_m),
        NA_real_
      ),
      reject = in_sig_family & !is.na(p_trimmed) & (p_trimmed <= q)
    ) %>%
    arrange(fam, p_sub)

  list(
    condition = cond_B,
    sig_major_fams = sig_fams,
    u1_cutoff = u1,
    major_table = major_df %>% mutate(adj = maj$adj),
    leaf_table = out
  )
}
```

```{r}
cluster_trim_list <- purrr::imap(spicy_sub_list, \(sp_sub, cc) {
  run_cluster_trim_one(
    spicy_sub   = sp_sub,
    spicy_major = spicy_major_list[[cc]],
    cond_B      = cc,
    q           = 0.05,
    major_method = "BY",   # <- or "BH"
    rho         = 0.5,
    m0_m        = 0.9,
    two_sided_z = FALSE
  )
})

cluster_trim_df <- bind_rows(lapply(cluster_trim_list, \(x) {
  x$leaf_table %>% mutate(condition = x$condition, u1_cutoff = x$u1_cutoff)
}))

# how many trimmed discoveries per condition?
cluster_trim_df %>%
  group_by(condition) %>%
  summarise(n_reject = sum(reject), .groups = "drop")
```

# Our Method

The method that we're proposing is instead of making to make our own internal p-values, we can make our own internal p-values. We can use this tree structure that we used for the treeBH example

- Level 0: root 
- Level 1: Major-Major interaction family
  - Group by the major cell types of the pair:
    - `T__T`
    - `T__B`
    - `T__NK`
    - `B__B`
    - `B__NK`
    - `NK__NK`
- Level 2: Subtype-Subtype within the damily
  - Under `T__T`
    - `CD4__CD4`
    - `CD4__CD8`
    - `CD4__Treg`
    - `CD8__CD8`
    - `CD8__Treg`
    - `Treg__Treg`
  - Under `T__B`
    - `CD4__B`
    - `CD8__B`
    - `Treg__B`
  - Under `B__B`
    - `B__B`
  - Under `B__NK`
    - `B__NK`
  - Under `NK__NK`
    - `NK__NK`
    

```{r}
make_tree_df_from_spicy <- function(spicy_major, spicy_sub, cond_B, eps = 1e-12) {
  colB <- paste0("condition", cond_B)

  stopifnot(colB %in% colnames(spicy_major$p.value))
  stopifnot(colB %in% colnames(spicy_sub$p.value))

  # ---- Level 1: major families
  maj_df <- tibble(
    interaction = rownames(spicy_major$p.value),
    node = canonical_pair(rownames(spicy_major$p.value)),
    level = 1L,
    parent = NA_character_,
    pvalue = as.numeric(spicy_major$p.value[, colB])
  ) %>%
    group_by(node) %>% slice(1) %>% ungroup() %>%   # drop duplicates after canonicalization
    mutate(pvalue = pmin(pmax(pvalue, eps), 1 - eps)) %>%
    select(node, level, parent, pvalue)

  # ---- Level 2: subtype leaves
  leaf_df <- tibble(
    interaction = rownames(spicy_sub$p.value),
    node = canonical_pair(rownames(spicy_sub$p.value)),
    level = 2L,
    pvalue = as.numeric(spicy_sub$p.value[, colB])
  ) %>%
    group_by(node) %>% slice(1) %>% ungroup() %>%   # drop duplicates after canonicalization
    tidyr::separate(node, into = c("a","b"), sep="__", remove = FALSE) %>%
    mutate(
      parent = canonical_pair(paste0(MAJOR_MAP[a], "__", MAJOR_MAP[b])),
      pvalue = pmin(pmax(pvalue, eps), 1 - eps)
    ) %>%
    select(node, level, parent, pvalue)

  # keep only leaves whose parent exists in major level
  leaf_df <- leaf_df %>% filter(parent %in% maj_df$node)

  bind_rows(maj_df, leaf_df) %>%
    arrange(level, parent, node)
}

run_levelwise_bh <- function(tree_df, q = 0.05) {
  L1 <- tree_df %>% filter(level == 1)
  L2 <- tree_df %>% filter(level == 2)

  # Level 1 BH across majors
  L1 <- L1 %>%
    mutate(tested = TRUE,
           reject = bh_reject(pvalue, q))

  sig_parents <- L1$node[L1$reject]

  # Level 2 BH within selected major families only
  L2 <- L2 %>%
    mutate(tested = parent %in% sig_parents,
           reject = FALSE)

  if (any(L2$tested)) {
    idx <- which(L2$tested)
    L2$reject[idx] <- bh_reject(L2$pvalue[idx], q)
  }

  bind_rows(L1, L2) %>% arrange(level, parent, node)
}

run_ours_one <- function(spicy_major, spicy_sub, cond_B, q = 0.05) {
  tree_df <- make_tree_df_from_spicy(spicy_major, spicy_sub, cond_B)

  res_ours <- run_levelwise_bh(tree_df, q = q)

  leaf_df <- tree_df %>% filter(level == 2)

  # vanilla BH across all leaves
  bh_all <- bh_reject(leaf_df$pvalue, q = q)
  bh_set <- leaf_df$node[bh_all]

  our_set <- res_ours %>% filter(level == 2, reject) %>% pull(node)

  list(
    tree_df = tree_df,
    res_ours = res_ours,
    sig_families = res_ours %>% filter(level == 1, reject),
    sig_leaves = res_ours %>% filter(level == 2, reject),
    compare = list(
      BH_all_n = length(bh_set),
      Our_n = length(our_set),
      overlap_n = length(intersect(bh_set, our_set)),
      ours_only = setdiff(our_set, bh_set),
      bh_only = setdiff(bh_set, our_set)
    )
  )
}
```

```{r}
ours_out <- purrr::imap(conds, \(cc, .i) {
  run_ours_one(
    spicy_major = spicy_major_list[[cc]],
    spicy_sub   = spicy_sub_list[[cc]],
    cond_B      = cc,
    q           = 0.05
  )
})
names(ours_out) <- conds

# Quick summary table
tibble(
  condition = conds,
  our_n = sapply(ours_out, \(x) nrow(x$sig_leaves)),
  bh_n  = sapply(ours_out, \(x) x$compare$BH_all_n),
  overlap = sapply(ours_out, \(x) x$compare$overlap_n)
)
```

# Summarize Everything


```{r}
# =========================================================
# 1) Define the ground-truth signal you injected (leaf-level)
# =========================================================

truth_map <- function(cond) {
  # return a list describing the injected signal
  # - leaves: canonical leaf names that are truly non-null
  # - scope: "leaf" if only one leaf is non-null; "internal" if a whole family is non-null
  if (cond == "LeafSignal_TtoNK") {
    return(list(scope = "leaf",     leaves = canonical_pair("CD4__NK")))
  }
  if (cond == "InternalSignal_TtoNK") {
    return(list(scope = "internal", leaves = canonical_pair(c("CD4__NK","CD8__NK","Treg__NK"))))
  }
  if (cond == "LeafSignal_NKtoCD4") {
    return(list(scope = "leaf",     leaves = canonical_pair("CD4__NK")))
  }
  if (cond == "InternalSignal_NKtoT") {
    return(list(scope = "internal", leaves = canonical_pair(c("CD4__NK","CD8__NK","Treg__NK"))))
  }

  # default (shouldn't happen if conds is as defined)
  list(scope = "none", leaves = character(0))
}

# =========================================================
# 2) Build a "base" leaf table per condition from spicy_sub
#    (pval/coef + fam mapping + truth columns)
# =========================================================

build_leaf_base <- function(spicy_sub, cond_B, eps = 1e-12) {
  # aligned + canonicalized + de-duped; already defined earlier by you
  df <- extract_leaf_stats(spicy_sub, conditionB = cond_B, eps = eps) %>%
    tidyr::separate(leaf, into = c("a","b"), sep="__", remove = FALSE) %>%
    mutate(
      fam = canonical_pair(paste0(MAJOR_MAP[a], "__", MAJOR_MAP[b]))
    ) %>%
    select(condition = interaction, leaf, fam, pval, coef) %>%
    mutate(condition = cond_B) %>%
    select(condition, fam, leaf, pval, coef) %>%
    arrange(fam, leaf)

  tr <- truth_map(cond_B)
  df %>%
    mutate(
      truth_signal = leaf %in% tr$leaves,
      truth_scope  = ifelse(truth_signal, tr$scope, "none")
    )
}

leaf_base_df <- purrr::imap_dfr(spicy_sub_list, \(sp_obj, cc) {
  build_leaf_base(sp_obj, cond_B = cc)
})

# =========================================================
# 3) Build method-specific "significance" tables at leaf level
# =========================================================

# --- BH across ALL leaves (per condition)
bh_all_df <- leaf_base_df %>%
  group_by(condition) %>%
  mutate(sig_BH_all = bh_reject(pval, q = 0.05)) %>%
  ungroup() %>%
  select(condition, leaf, sig_BH_all)

# --- TreeBH (you already made treebh_sub_df)
treebh_sig_df <- treebh_sub_df %>%
  transmute(condition, leaf = leaf, sig_TreeBH = reject_treebh)

# --- DART (you already made dart_df)
dart_sig_df <- dart_df %>%
  transmute(condition, leaf, sig_DART = reject)

# --- PFilter (you already made pfilter_df)
pfilter_sig_df <- pfilter_df %>%
  transmute(condition, leaf, sig_PFilter = reject)

# --- Cluster+Trim (you already made cluster_trim_df)
clustertrim_sig_df <- cluster_trim_df %>%
  transmute(condition, leaf, sig_ClusterTrim = reject)

# --- Our method (ours_out list: each has $sig_leaves with node names)
ours_sig_df <- purrr::imap_dfr(ours_out, \(x, cc) {
  tibble(condition = cc, leaf = x$sig_leaves$node, sig_Ours = TRUE)
})

# =========================================================
# 4) Join everything into ONE summary table
# =========================================================

summary_leaf_df <- leaf_base_df %>%
  left_join(bh_all_df,         by = c("condition","leaf")) %>%
  left_join(treebh_sig_df,     by = c("condition","leaf")) %>%
  left_join(dart_sig_df,       by = c("condition","leaf")) %>%
  left_join(ours_sig_df,       by = c("condition","leaf")) %>%
  left_join(pfilter_sig_df,    by = c("condition","leaf")) %>%
  left_join(clustertrim_sig_df,by = c("condition","leaf")) %>%
  mutate(
    sig_TreeBH      = tidyr::replace_na(sig_TreeBH, FALSE),
    sig_DART        = tidyr::replace_na(sig_DART, FALSE),
    sig_Ours        = tidyr::replace_na(sig_Ours, FALSE),
    sig_PFilter     = tidyr::replace_na(sig_PFilter, FALSE),
    sig_ClusterTrim = tidyr::replace_na(sig_ClusterTrim, FALSE)
  ) %>%
  arrange(condition, fam, leaf, pval)

summary_leaf_df

# Optionally: write to CSV
# write.csv(summary_leaf_df, "summary_leaf_table.csv", row.names = FALSE)

# =========================================================
# 5) OPTIONAL: per-condition confusion summary per method
# =========================================================

confusion_one <- function(df, sig_col) {
  sig <- df[[sig_col]]
  truth <- df$truth_signal
  tibble(
    method = sig_col,
    TP = sum(sig & truth),
    FP = sum(sig & !truth),
    FN = sum(!sig & truth),
    TN = sum(!sig & !truth)
  )
}

perf_by_condition <- summary_leaf_df %>%
  group_by(condition) %>%
  group_modify(\(d, ...) {
    bind_rows(
      confusion_one(d, "sig_BH_all"),
      confusion_one(d, "sig_TreeBH"),
      confusion_one(d, "sig_DART"),
      confusion_one(d, "sig_Ours"),
      confusion_one(d, "sig_PFilter"),
      confusion_one(d, "sig_ClusterTrim")
    )
  }) %>%
  ungroup()

perf_by_condition
```

```{r}
summary_leaf_df %>% filter(truth_scope == "leaf")
summary_leaf_df %>% filter(truth_scope == "internal")
```

```{r}

df_filtered <- summary_leaf_df %>%
  filter(
    truth_signal == TRUE |
    (
      truth_signal == FALSE &
      (
        sig_BH_all |
        sig_TreeBH |
        sig_DART |
        sig_Ours |
        sig_PFilter |
        sig_ClusterTrim
      )
    )
  )

view(df_filtered)
```

